---
title: "Bayesian Modeling of STIR Effects on Water Quality"
author: "AJ Brown"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE
)

# Core libs
library(rethinking)
library(posterior)
library(cmdstanr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(readr)
library(stringr)
library(purrr)
```


```{r setup, include=FALSE}
# If you rely on helper functions (e.g., make_year_dist_mat), source them here:
source("./stir-bayes-backend.R")
```

# Model 1.7 (Concentration + Volume Joint Model with Crop and Residue Effects with latent true processes)

* Joint Bayesian model for **OUT concentration and OUT volume** fit simultaneously
* Model uses multivariate normal (MVN) hierarchical structure for analyte-specific parameters
* Model is parametrized in **non-centered** form for better sampling efficiency
* **Analyte-specific** intercepts, STIR slopes, and inflow-concentration slopes
* **Volume** modeled as a function of STIR, with Bayesian imputation for missing VOL
* **Inflow concentration (CIN)** now handled via Bayesian imputation (`CIN ~ dnorm(0,1)`), not zero-substitution
* Includes all DAG-identified adjustment variables affecting concentration
  * { Analyte, Block, Inflow Conc., Irrigation Count, Outflow vol, Sampler Method, duplicate status, flume type }
* Full uncertainty propagation from CIN, VOL, STIR into analyte-specific concentration predictions
* Supports **scenario-based simulation** for STIR → Load via generative functions (conditional or averaged scenarios)
* Incorporates multi-output Gaussian Process (GP) on Year to capture temporal correlation in residuals for each analyte
* v1.5 only has cmdstan model, no rethinking/ulam version
* v1.6 adds crop type effects and residue cover at planting effects
* v1.7 promotes residue, volume, and inflow concentration to latent processes with full Bayesian imputation and joint propagation. Also uses new python wq_cleaned.csv, which has been tested to be identical to previous versions.


# 1. Load and lightly type-enforce the Bayes-clean dataset (from Python pipeline)

```{r load-clean}
# Prefer the Python-produced Bayes-clean dataset
clean_path <- "out/wq_cleaned.csv"
if (!file.exists(clean_path)) {
  alt <- file.path("..", clean_path)
  if (file.exists(alt)) {
    clean_path <- alt
  } else {
    stop(
      "Bayes-clean file not found. Expected 'out/wq_cleaned.csv'.
",
      "Run the Python pipeline first:
",
      "  python code/run_pipeline.py
"
    )
  }
}

wq_clean <- readr::read_csv(clean_path, show_col_types = FALSE)

abbr_levels <- c("NH4","ICP","NO3","NOx","NO2","NPOC","OP","Se","TDS","TKN","TN","TP","TSP","TSS")

# Defensive type enforcement (lightweight, no re-cleaning)
wq_clean <- wq_clean %>%
  dplyr::mutate(
    # Dates
    Date        = if ("Date" %in% names(.)) as.Date(Date) else Date,
    PlantDate   = if ("PlantDate" %in% names(.)) as.Date(PlantDate) else PlantDate,
    HarvestDate = if ("HarvestDate" %in% names(.)) as.Date(HarvestDate) else HarvestDate,

    # Numeric/integer fields commonly used downstream
    Year        = if ("Year" %in% names(.)) as.integer(Year) else Year,
    SeasonYear  = if ("SeasonYear" %in% names(.)) as.integer(SeasonYear) else SeasonYear,

    # Factors/categoricals often used downstream
    Treatment    = if ("Treatment" %in% names(.)) factor(Treatment, levels = c("CT","MT","ST")) else Treatment,
    Rep          = if ("Rep" %in% names(.)) factor(Rep) else Rep,
    Irrigation   = if ("Irrigation" %in% names(.)) factor(Irrigation) else Irrigation,
    SampleMethod = if ("SampleMethod" %in% names(.)) factor(SampleMethod) else SampleMethod,
    FlumeMethod  = if ("FlumeMethod" %in% names(.)) factor(FlumeMethod) else FlumeMethod,
    Duplicate    = if ("Duplicate" %in% names(.)) as.integer(Duplicate) else Duplicate,
    Analyte      = if ("Analyte" %in% names(.)) factor(Analyte) else Analyte,
    analyte_abbr = if ("analyte_abbr" %in% names(.)) factor(analyte_abbr, levels = abbr_levels) else analyte_abbr,
    Crop         = if ("Crop" %in% names(.)) factor(Crop) else Crop,

    # Add original row number for tracking
    orig_row = dplyr::row_number()
  )

# Sanity checks (quick)
req_cols <- c("cout_z","cin_z","stir_season_z","volume_z","Treatment","Rep","SampleMethod","FlumeMethod","Irrigation","Duplicate","Crop","analyte_abbr")
missing_cols <- setdiff(req_cols, names(wq_clean))
if (length(missing_cols) > 0) {
  warning("Missing expected columns in wq_cleaned.csv: ", paste(missing_cols, collapse = ", "))
}

cat("Loaded Bayes-clean dataset:", clean_path, "
")
cat("Rows:", nrow(wq_clean), " | Columns:", ncol(wq_clean), "
")

dplyr::glimpse(wq_clean)
```

---

# 2. Prepare model-ready dataset

Select analytes for modeling:

```{r subset-analytes}
# analytes_keep <- c(
#   "OP", "TP", "ICP", "NO3", "TN", "TSS",
#   "TKN", "TSP", "NH4", "NPOC", "Se", "NO2",
#   "NOx", "TDS"
# )

analytes_keep <- c(
  "OP", "TP", "NO3", "TN", "TSS",
  "TKN", "NH4", "Se", "NO2", "TDS"
)

d_mod <- wq_clean %>%
  dplyr::filter(analyte_abbr %in% analytes_keep) %>%
  dplyr::mutate(mod_row = dplyr::row_number()) %>%
  droplevels()

na_table <- d_mod %>%
  dplyr::group_by(analyte_abbr) %>%
  dplyr::summarize(
    n_total   = dplyr::n(),
    n_value   = sum(!is.na(Result_mg_L)),
    n_na      = sum(is.na(Result_mg_L)),
    prop_na   = n_na / n_total, #proportion missing
    .groups   = "drop"
  )

na_table

```

Create analyte index and build data list for ulam:

```{r make-stan-dat}
# 2. Build indices and standardized predictors -----------------------------

# Choose residue metric here:
# - "cover": uses residue proportion in (0,1) (compatible with current Stan logit-residue model)
# - "drymass": uses z-scored log(drymass_kg_m2) (requires Stan residue model update)
residue_metric <- "cover"  # "cover" or "drymass"

d_mod <- d_mod %>%
  mutate(
    # analyte index
    A = as.integer(analyte_abbr),

    # Block (Rep) index
    B = as.integer(Rep),

    # Sampler method index
    S = as.integer(SampleMethod),

    # Flume type index
    Fu = as.integer(FlumeMethod),

    # Irrigation count as numeric
    irr_num = dplyr::case_when(
      # Explicitly drop storm events
      as.character(Irrigation) %in% c("S1", "S2") ~ NA_integer_,

      # Valid numeric irrigation counts
      TRUE ~ as.integer(readr::parse_number(as.character(Irrigation)))
    ),

    # Inflow concentration (standardized), keep NAs for Bayesian imputation
    cin_std = cin_z,

    # Duplicate status as 0/1
    dup = as.integer(Duplicate),

    # Year index
    Y = as.integer(as.factor(Year)),

    # Crop type index (categorical)
    Cr = as.integer(Crop),

    # -----------------------------------------------------------------------
    # Residue percent cover -> proportion in (0,1) with epsilon adjustment
    # -----------------------------------------------------------------------
    residue_prop = {
      x <- as.numeric(.data$Residue_PercentCover)

      # Convert 0–100 to 0–1 if needed
      x <- ifelse(is.na(x), NA_real_,
                  ifelse(x > 1, x / 100, x))

      # Epsilon adjustment for Stan (strictly inside 0,1)
      eps <- 1e-4
      ifelse(is.na(x), NA_real_,
             pmin(pmax(x, eps), 1 - eps))
    },

    # -----------------------------------------------------------------------
    # Residue dry mass (kg/m^2) -> log transform (<=0 becomes NA)
    # -----------------------------------------------------------------------
    drymass_kg_m2 = as.numeric(.data$Residue_DryMass_kg_m2),
    drymass_log = dplyr::if_else(
      is.na(drymass_kg_m2) | drymass_kg_m2 <= 0,
      NA_real_,
      log(drymass_kg_m2)
    )
  )

# Drop storm rows (no ordinal irrigation meaning)
d_mod <- d_mod %>%
  dplyr::filter(!is.na(irr_num))

# Z-score drymass_log using observed values only (keeps NA as NA)
drymass_mu <- mean(d_mod$drymass_log, na.rm = TRUE)
drymass_sd <- sd(d_mod$drymass_log, na.rm = TRUE)

d_mod <- d_mod %>%
  mutate(
    drymass_log_z = (drymass_log - drymass_mu) / drymass_sd,

    # Single residue input used by Stan
    RES_input = dplyr::case_when(
      residue_metric == "cover"   ~ residue_prop,
      residue_metric == "drymass" ~ drymass_log_z,
      TRUE ~ residue_prop
    )
  )

# instantiate distance matrix for gaussian process on year
years_used <- sort(unique(d_mod$Year))
mat <- make_year_dist_mat(years_used)

# lookup code for later
analyte_lookup <- d_mod %>%
  distinct(A, analyte_abbr) %>%
  arrange(A)

block_lookup <- d_mod %>%
  distinct(B, Rep) %>%
  arrange(B)

sampler_lookup <- d_mod %>%
  distinct(S, SampleMethod) %>%
  arrange(S)

flume_lookup <- d_mod %>%
  distinct(Fu, FlumeMethod) %>%
  arrange(Fu)

irrigation_lookup <- d_mod %>%
  distinct(irr_num, Irrigation) %>%
  arrange(irr_num)

duplicate_lookup <- d_mod %>%
  distinct(dup, Duplicate) %>%
  arrange(dup)

crop_lookup <- d_mod %>%
  distinct(Cr, Crop) %>%
  arrange(Cr)

# prepare data list for stan
stan_dat <- list(
  N    = nrow(d_mod),

  # group counts
  A_n  = length(unique(d_mod$A)),
  B_n  = length(unique(d_mod$B)),
  S_n  = length(unique(d_mod$S)),
  F_n  = length(unique(d_mod$Fu)),
  D_n  = length(unique(d_mod$dup)),
  Y_n  = length(unique(d_mod$Y)),
  Cr_n = length(unique(d_mod$Cr)),

  # indices
  A    = d_mod$A,
  B    = d_mod$B,
  S    = d_mod$S,
  Fu   = d_mod$Fu,
  Y    = d_mod$Y,
  Cr   = d_mod$Cr,

  # distance matrix for year effects
  D    = mat,

  # outcome
  C    = d_mod$cout_z,

  # main exposure
  STIR = d_mod$stir_season_z,

  # other covariates from DAG
  CIN  = d_mod$cin_std,
  VOL  = d_mod$volume_z,
  IRR  = d_mod$irr_num,
  DUP  = d_mod$dup,

  # residue covariate (cover or drymass, depending on residue_metric)
  RES  = d_mod$RES_input
)

# Optional: quick guardrails
if (identical(residue_metric, "cover")) {
  stopifnot(all(is.na(stan_dat$RES) | (stan_dat$RES > 0 & stan_dat$RES < 1)))
}


```

```{r check-na-vars}
# Identify the stan_dat entries that are vectors of length N (i.e., row-level variables)
vec_names <- names(stan_dat)[sapply(stan_dat, function(x) length(x) == stan_dat$N)]

# Loop through these and report NAs
for (nm in vec_names) {
  vals <- stan_dat[[nm]]
  na_idx <- which(is.na(vals))

  cat("\n==============================\n")
  cat("Variable:", nm, "\n")

  if (length(na_idx) == 0) {
    cat("No NA values detected.\n")
  } else {
    cat("NA count:", length(na_idx), "\n")
    cat("Rows with NA:", paste(head(na_idx, 25), collapse = ", "),
        if (length(na_idx) > 25) "..." else "", "\n\n")

    # Show the corresponding d_mod rows
    cat("Example offending rows from d_mod:\n")
    print(d_mod[na_idx[1:min(10, length(na_idx))], ])
  }
}

```


# 3. Hierarchical Bayesian model

```{r cmdstan model fit}
# Fit cmdstan model: m_stir_mogp_v1p7p5.stan
# Key requirement: Stan cannot accept NA in data vectors, so we:
#   (1) record missing row indices (missidx),
#   (2) fill missing entries with harmless placeholders,
#   (3) pass missidx + N_miss so the model can exclude those rows from the observation likelihood
#       (and impute via latent parameters where applicable).
#
# For residue specifically, the Stan model expects a PROPORTION strictly inside (0,1),
# so we clamp observed values with epsilon and fill missing with 0.1.

# Force recompile to ensure v1p7p5 is actually what runs
mod <- cmdstan_model("m_stir_mogp_v1p7p5.stan", force_recompile = TRUE)

# Start from the base stan_dat you built earlier
stan_dat_cmd <- stan_dat

# -------------------------------------------------------------------
# 1) Pull row-level vectors (must all be length N)
# -------------------------------------------------------------------
C_vec   <- d_mod$cout_z
CIN_vec <- d_mod$cin_std
VOL_vec <- d_mod$volume_z

# Crop index used by Stan (must be integer-coded 1..Cr_n already)
Cr_vec  <- d_mod$Cr

# Residue observations:
# Prefer precomputed 0–1 proportion column if present; otherwise derive from percent cover.
if ("residue_prop" %in% names(d_mod)) {
  RES_vec_raw <- as.numeric(d_mod$residue_prop)
} else if ("Residue_PercentCover" %in% names(d_mod)) {
  RES_vec_raw <- as.numeric(d_mod$Residue_PercentCover)
  # Convert 0–100 to 0–1 if needed (if already 0–1, leave as is)
  RES_vec_raw <- ifelse(is.na(RES_vec_raw), NA_real_,
                        ifelse(RES_vec_raw > 1, RES_vec_raw / 100, RES_vec_raw))
} else {
  stop("No residue column found. Expected 'residue_prop' or 'Residue_PercentCover'.")
}

stopifnot(
  length(C_vec)       == stan_dat_cmd$N,
  length(CIN_vec)     == stan_dat_cmd$N,
  length(VOL_vec)     == stan_dat_cmd$N,
  length(Cr_vec)      == stan_dat_cmd$N,
  length(RES_vec_raw) == stan_dat_cmd$N
)

# -------------------------------------------------------------------
# 2) Helpers: build missidx + NA-free vectors for Stan
# -------------------------------------------------------------------
prep_miss_gaussian <- function(x, fill = 0) {
  missidx <- which(is.na(x))   # 1-based indices (Stan convention)
  x_clean <- as.numeric(x)
  if (length(missidx) > 0) x_clean[missidx] <- fill
  list(
    missidx = missidx,
    N_miss  = length(missidx),
    clean   = x_clean
  )
}

prep_miss_residue <- function(x, eps = 1e-6, fill = 0.1) {
  x_num <- as.numeric(x)

  missidx <- which(is.na(x_num))
  obs_idx <- setdiff(seq_along(x_num), missidx)

  if (length(obs_idx) > 0) {
    bad_nonfinite <- obs_idx[!is.finite(x_num[obs_idx])]
    if (length(bad_nonfinite) > 0) {
      stop("RES contains non-finite values at rows: ",
           paste(head(bad_nonfinite, 20), collapse = ", "),
           if (length(bad_nonfinite) > 20) " ..." else "")
    }
  }

  if (length(obs_idx) > 0 && any(x_num[obs_idx] > 1, na.rm = TRUE)) {
    x_num[obs_idx] <- x_num[obs_idx] / 100
  }

  if (length(obs_idx) > 0) {
    bad_range <- obs_idx[x_num[obs_idx] < 0 | x_num[obs_idx] > 1]
    if (length(bad_range) > 0) {
      stop("RES has values outside [0,1] at rows: ",
           paste(head(bad_range, 20), collapse = ", "),
           if (length(bad_range) > 20) " ..." else "",
           "\nCheck residue scaling upstream (should be a 0–1 proportion).")
    }
  }

  if (length(obs_idx) > 0) {
    x_num[obs_idx] <- pmin(pmax(x_num[obs_idx], eps), 1 - eps)
  }

  if (length(missidx) > 0) x_num[missidx] <- fill

  list(
    missidx = missidx,
    N_miss  = length(missidx),
    clean   = x_num
  )
}

# -------------------------------------------------------------------
# 3) Apply helpers
# -------------------------------------------------------------------
imp_C    <- prep_miss_gaussian(C_vec,   fill = 0)
imp_CIN  <- prep_miss_gaussian(CIN_vec, fill = 0)
imp_VOL  <- prep_miss_gaussian(VOL_vec, fill = 0)
imp_RES  <- prep_miss_residue(RES_vec_raw, eps = 1e-4, fill = 0.5)

# -------------------------------------------------------------------
# 4) Populate cmdstan data list (no NA allowed in vectors)
# -------------------------------------------------------------------
stan_dat_cmd$C         <- imp_C$clean
stan_dat_cmd$N_C_miss  <- imp_C$N_miss
stan_dat_cmd$C_missidx <- if (imp_C$N_miss == 0) integer(0) else as.array(as.integer(imp_C$missidx))

stan_dat_cmd$CIN         <- imp_CIN$clean
stan_dat_cmd$N_CIN_miss  <- imp_CIN$N_miss
stan_dat_cmd$CIN_missidx <- if (imp_CIN$N_miss == 0) integer(0) else as.array(as.integer(imp_CIN$missidx))

stan_dat_cmd$VOL         <- imp_VOL$clean
stan_dat_cmd$N_VOL_miss  <- imp_VOL$N_miss
stan_dat_cmd$VOL_missidx <- if (imp_VOL$N_miss == 0) integer(0) else as.array(as.integer(imp_VOL$missidx))

stan_dat_cmd$RES         <- imp_RES$clean
stan_dat_cmd$N_RES_miss  <- imp_RES$N_miss
stan_dat_cmd$RES_missidx <- if (imp_RES$N_miss == 0) integer(0) else as.array(as.integer(imp_RES$missidx))

# Explicit missingness flags (guarantees consistency with current cleaned vectors)
stan_dat_cmd$is_C_miss   <- as.array(as.integer(is.na(C_vec)))
stan_dat_cmd$is_VOL_miss <- as.array(as.integer(is.na(VOL_vec)))
stan_dat_cmd$is_RES_miss <- as.array(as.integer(is.na(RES_vec_raw)))

# Crop index (already integer-coded in d_mod)
Cr_n_now <- length(unique(Cr_vec))
stan_dat_cmd$Cr   <- as.array(as.integer(Cr_vec))
if (is.null(stan_dat_cmd$Cr_n) || stan_dat_cmd$Cr_n != Cr_n_now) {
  stan_dat_cmd$Cr_n <- Cr_n_now
}

# -------------------------------------------------------------------
# 5) Consistency checks + diagnostics
# -------------------------------------------------------------------
stopifnot(
  all(is.finite(stan_dat_cmd$C)),
  all(is.finite(stan_dat_cmd$CIN)),
  all(is.finite(stan_dat_cmd$VOL)),
  all(is.finite(stan_dat_cmd$RES)),
  all(stan_dat_cmd$RES > 0),
  all(stan_dat_cmd$RES < 1),
  isTRUE(all(stan_dat_cmd$Cr >= 1L)),
  isTRUE(all(stan_dat_cmd$Cr <= stan_dat_cmd$Cr_n))
)

cat("\n[INFO] Missing counts:\n")
cat("  N_C_miss    =", stan_dat_cmd$N_C_miss, "\n")
cat("  N_CIN_miss  =", stan_dat_cmd$N_CIN_miss, "\n")
cat("  N_VOL_miss  =", stan_dat_cmd$N_VOL_miss, "\n")
cat("  N_RES_miss  =", stan_dat_cmd$N_RES_miss, "\n")
cat("[INFO] Crop levels (Cr_n) =", stan_dat_cmd$Cr_n, "\n")

cat("[INFO] Residue summary (raw, before clamp/fill):\n")
print(summary(RES_vec_raw))
cat("[INFO] Residue summary (Stan-ready, after clamp/fill):\n")
print(summary(stan_dat_cmd$RES))

# -------------------------------------------------------------------
# 6) Sample with cmdstanr
# -------------------------------------------------------------------
dir.create("out_cmdstanr", showWarnings = FALSE)

fit <- mod$sample(
  data            = stan_dat_cmd,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 2000,
  iter_sampling   = 1000,
  refresh         = 100,
  show_messages   = TRUE,
  output_dir      = "out_cmdstanr",
  save_warmup     = TRUE,
  adapt_delta     = 0.95,
  max_treedepth   = 15
)

```


```{r cmdstan model fit}
saveRDS(fit, file = "./out_cmdstanr/fit_mogp_v1p7p5.rds")
# fit <- readRDS("./out_cmdstanr/fit_mogp_v1p7p5.rds")

```


---

# 4. Posterior inspection

```{r stan posterior summary}
summarize_cmdstan <- function(fit, vars = NULL, n_show = 30) {
  # cmdstanr summary
  s <- if (is.null(vars)) {
    fit$summary()
  } else {
    fit$summary(variables = vars)
  }

  # basic display
  cat("\n=== Top", n_show, "rows of cmdstan summary ===\n")
  print(head(s, n_show), row.names = FALSE)

  # diagnostics (guard against missing columns depending on cmdstanr version)
  rhat_vals <- if ("rhat" %in% names(s)) s$rhat else rep(NA_real_, nrow(s))
  ess_vals  <- if ("ess_bulk" %in% names(s)) s$ess_bulk else rep(NA_real_, nrow(s))

  # helper to avoid which.min/max on all-NA
  safe_which_min <- function(x) if (all(is.na(x))) NA_integer_ else which.min(x)
  safe_which_max <- function(x) if (all(is.na(x))) NA_integer_ else which.max(x)

  i_min_rhat <- safe_which_min(rhat_vals)
  i_max_rhat <- safe_which_max(rhat_vals)
  i_min_ess  <- safe_which_min(ess_vals)
  i_max_ess  <- safe_which_max(ess_vals)

  cat("\n=== Rhat Diagnostics ===\n")
  cat("Min Rhat:   ", suppressWarnings(min(rhat_vals, na.rm = TRUE)),
      if (!is.na(i_min_rhat)) paste0(" (", s$variable[i_min_rhat], ")") else "", "\n")
  cat("Median Rhat:", suppressWarnings(median(rhat_vals, na.rm = TRUE)), "\n")
  cat("Max Rhat:   ", suppressWarnings(max(rhat_vals, na.rm = TRUE)),
      if (!is.na(i_max_rhat)) paste0(" (", s$variable[i_max_rhat], ")") else "", "\n")

  cat("\n=== Bulk ESS Diagnostics ===\n")
  cat("Min ESS:    ", suppressWarnings(min(ess_vals, na.rm = TRUE)),
      if (!is.na(i_min_ess)) paste0(" (", s$variable[i_min_ess], ")") else "", "\n")
  cat("Median ESS: ", suppressWarnings(median(ess_vals, na.rm = TRUE)), "\n")
  cat("Max ESS:    ", suppressWarnings(max(ess_vals, na.rm = TRUE)),
      if (!is.na(i_max_ess)) paste0(" (", s$variable[i_max_ess], ")") else "", "\n")

  # v1p7p5 convenience: highlight crop/residue blocks if present
  # Names must match the Stan parameter names exactly
  new_blocks <- c("beta_res_V", "beta_res_C", "gamma_Cr_V", "gamma_Cr")
  
  # cmdstanr summary uses rownames in $variable like:
  #  beta_res_V
  #  beta_res_C[1]
  #  gamma_Cr_V[3]
  #  gamma_Cr[2,5]
  is_in_block <- function(v, base) grepl(paste0("^", base, "(\\[|$)"), v)
  
  sel <- Reduce(
    `|`,
    lapply(new_blocks, function(b) is_in_block(s$variable, b))
  )
  
  if (any(sel)) {
    cat("\n=== v1p7p5: Crop/Residue terms present in fit (first 50 rows) ===\n")
    print(utils::head(s[sel, ], 50), row.names = FALSE)
  
    # quick counts by block (helps confirm dimensions at a glance)
    cat("\n=== v1p7p5: Term counts by parameter block ===\n")
    for (b in new_blocks) {
      n_b <- sum(is_in_block(s$variable, b))
      if (n_b > 0) cat(sprintf("%-12s : %d\n", b, n_b))
    }
  } else {
    cat("\n[WARN] v1p7p5: No crop/residue parameters detected in fit summary.\n")
    cat("[WARN] Expected blocks: beta_res_V, beta_res_C[...], gamma_Cr_V[...], gamma_Cr[...,...]\n")
  }

  invisible(s)
}

s_all <- summarize_cmdstan(fit, n_show = 50)

```

```{r}
vars_important <- c(
  "a_V","b_V","beta_res_V",
  "beta_vol","mu_beta_irr", "beta_irr","beta_res_C",
  "alpha","beta_stir","beta_cin","beta_dup",
  "sigma_V","sigma_VOL_obs","sigma_analyte","sigma_C_obs",
  "res_base","b_res_stir","gamma_Cr_res","sigma_Cr_res","sigma_res_obs",
  "etasq_year","rhosq_year"
)

vars_betas <- c(
  "alpha","beta_stir","beta_cin","beta_dup","beta_res_C","beta_irr",
  "gamma_Cr","gamma_B","gamma_S","gamma_F"
)

vars_sigma <- c(
  "mu_log_sigma_C_obs",
  "tau_log_sigma_C_obs",
  paste0("z_log_sigma_C_obs[", 1:stan_dat_cmd$A_n, "]"),
  "sigma_analyte",
  "sigma_V",
  "sigma_VOL_obs",
  "sigma_res_obs",
  "sigma_Cr_res"
)

vars_irr_mvn <- c(
  "mu_beta_irr",
  paste0("sigma_beta_irr[", 1:stan_dat_cmd$A_n, "]"),
  paste0("z_beta_irr[", 1:stan_dat_cmd$A_n, "]"),
  "L_beta_irr"
)

print(fit$summary(variables = vars_important), n = 100)
print(fit$summary(variables = vars_betas), n = 300)
print(fit$summary(variables = vars_sigma), n = 100)
print(fit$summary(variables = vars_irr_mvn), n = 200)
```


```{r}
# Takes a long time to run (2-3+ hours)
#cmdstan_dashboard(fit, vars = vars_important, trank = TRUE)
cmdstan_dashboard(fit)

```


```{r stan analyte effects}
## =====================================================================
## Extract posterior samples from cmdstanr into a "rethinking-style" list
## v1p7p5: analyte × crop effects for concentration + residue-safe outputs
## DROP-IN REPLACEMENT: updated for v1p7p5 sigma structure:
##   - sigma_analyte is GLOBAL (scalar) process SD for C_true
##   - sigma_C_obs[A] is analyte-specific OBS SD for C (hierarchical on log scale)
## =====================================================================

dat_used <- if (exists("stan_dat_cmd")) stan_dat_cmd else stan_dat

# Dimensions
A_n  <- dat_used$A_n
B_n  <- dat_used$B_n
S_n  <- dat_used$S_n
F_n  <- dat_used$F_n
Y_n  <- dat_used$Y_n
Cr_n <- dat_used$Cr_n

# Residue placeholder check (all zeros means no supported effect yet)
res_sd <- sd(dat_used$RES, na.rm = TRUE)
res_is_placeholder <- is.na(res_sd) || res_sd == 0

# Draws matrix
draws_mat <- fit$draws(
  variables = c(
    # analyte-level fixed effects
    "alpha", "beta_stir", "beta_cin", "beta_dup",

    # random effects
    "gamma_B", "gamma_S", "gamma_F",

    # volume model terms
    "beta_vol", "beta_irr", "a_V", "b_V", "sigma_V", "sigma_VOL_obs",

    # GP on year
    "F_year", "etasq_year", "rhosq_year",

    # residue as predictor in volume + concentration
    "beta_res_V", "beta_res_C",

    # crop effects (v1p7p5)
    "gamma_Cr", "gamma_Cr_V",
    "sigma_Cr", "sigma_Cr_V",

    # concentration SD structure (v1p7p5)
    "sigma_analyte",
    "sigma_C_obs",
    "mu_log_sigma_C_obs", "tau_log_sigma_C_obs", "z_log_sigma_C_obs",

    # residue submodel (useful for checking + reporting)
    "res_base", "b_res_stir", "gamma_Cr_res", "sigma_Cr_res", "sigma_res_obs"
  ),
  format = "matrix"
)

n_draws <- nrow(draws_mat)

get_param_mat <- function(draws_mat, par) {
  sel <- grepl(paste0("^", par, "\\["), colnames(draws_mat))
  if (!any(sel)) stop("No indexed columns found for parameter '", par, "'.")
  draws_mat[, sel, drop = FALSE]
}

get_param_vec <- function(draws_mat, par) {
  if (!(par %in% colnames(draws_mat))) stop("No scalar column found for parameter '", par, "'.")
  draws_mat[, par]
}

# ---- analyte-level matrices (draws × A_n)
alpha_mat      <- get_param_mat(draws_mat, "alpha")
beta_stir_mat  <- get_param_mat(draws_mat, "beta_stir")
beta_cin_mat   <- get_param_mat(draws_mat, "beta_cin")
beta_dup_mat   <- get_param_mat(draws_mat, "beta_dup")
beta_res_C_mat <- get_param_mat(draws_mat, "beta_res_C")

# ---- analyte × block random effects: draws × A_n × B_n
gamma_B_mat <- get_param_mat(draws_mat, "gamma_B")
gamma_B_arr <- array(NA_real_, dim = c(n_draws, A_n, B_n))
k <- 1L
for (b in seq_len(B_n)) {
  for (a in seq_len(A_n)) {
    gamma_B_arr[, a, b] <- gamma_B_mat[, k]; k <- k + 1L
  }
}

# ---- analyte × sampler random effects: draws × A_n × S_n
gamma_S_mat <- get_param_mat(draws_mat, "gamma_S")
gamma_S_arr <- array(NA_real_, dim = c(n_draws, A_n, S_n))
k <- 1L
for (s in seq_len(S_n)) {
  for (a in seq_len(A_n)) {
    gamma_S_arr[, a, s] <- gamma_S_mat[, k]; k <- k + 1L
  }
}

# ---- analyte × flume random effects: draws × A_n × F_n
gamma_F_mat <- get_param_mat(draws_mat, "gamma_F")
gamma_F_arr <- array(NA_real_, dim = c(n_draws, A_n, F_n))
k <- 1L
for (f in seq_len(F_n)) {
  for (a in seq_len(A_n)) {
    gamma_F_arr[, a, f] <- gamma_F_mat[, k]; k <- k + 1L
  }
}

# ---- GP year effects: draws × Y_n × A_n
F_mat <- get_param_mat(draws_mat, "F_year")
F_array <- array(NA_real_, dim = c(n_draws, Y_n, A_n))
k <- 1L
for (a in seq_len(A_n)) {
  for (y in seq_len(Y_n)) {
    F_array[, y, a] <- F_mat[, k]; k <- k + 1L
  }
}

# ---- crop effects on concentration: gamma_Cr[A, Cr]
gamma_Cr_mat <- get_param_mat(draws_mat, "gamma_Cr")
if (ncol(gamma_Cr_mat) != A_n * Cr_n) {
  stop("gamma_Cr dimension mismatch: expected ", A_n * Cr_n,
       " columns (A_n * Cr_n) but found ", ncol(gamma_Cr_mat), ".")
}
gamma_Cr_arr <- array(NA_real_, dim = c(n_draws, A_n, Cr_n))
k <- 1L
for (cr in seq_len(Cr_n)) {
  for (a in seq_len(A_n)) {
    gamma_Cr_arr[, a, cr] <- gamma_Cr_mat[, k]; k <- k + 1L
  }
}

# ---- crop effects on volume: gamma_Cr_V[Cr]
gamma_Cr_V_mat <- get_param_mat(draws_mat, "gamma_Cr_V")
if (ncol(gamma_Cr_V_mat) != Cr_n) {
  stop("gamma_Cr_V dimension mismatch: expected ", Cr_n,
       " columns (Cr_n) but found ", ncol(gamma_Cr_V_mat), ".")
}

# ---- scalars
beta_vol_vec    <- get_param_vec(draws_mat, "beta_vol")
beta_irr_mat    <- get_param_mat(draws_mat, "beta_irr")
if (ncol(beta_irr_mat) != A_n) {
  stop("beta_irr dimension mismatch: expected ", A_n,
       " columns (A_n) but found ", ncol(beta_irr_mat), ".")
}
a_V_vec         <- get_param_vec(draws_mat, "a_V")
b_V_vec         <- get_param_vec(draws_mat, "b_V")
sigma_V_vec     <- get_param_vec(draws_mat, "sigma_V")
sigma_VOL_obs_vec <- get_param_vec(draws_mat, "sigma_VOL_obs")
beta_res_V_vec  <- get_param_vec(draws_mat, "beta_res_V")

# ---- sigma objects (v1p7p5)
sigma_analyte_vec <- get_param_vec(draws_mat, "sigma_analyte")   # GLOBAL process SD for C_true

sigma_C_obs_mat <- get_param_mat(draws_mat, "sigma_C_obs")       # draws × A_n (measurement SDs)
mu_log_sigma_C_obs_vec  <- get_param_vec(draws_mat, "mu_log_sigma_C_obs")
tau_log_sigma_C_obs_vec <- get_param_vec(draws_mat, "tau_log_sigma_C_obs")
z_log_sigma_C_obs_mat   <- get_param_mat(draws_mat, "z_log_sigma_C_obs")

sigma_Cr_mat   <- get_param_mat(draws_mat, "sigma_Cr")           # draws × A_n (crop SD per analyte)
sigma_Cr_V_vec <- get_param_vec(draws_mat, "sigma_Cr_V")         # scalar

etasq_year_vec  <- get_param_vec(draws_mat, "etasq_year")
rhosq_year_vec  <- get_param_vec(draws_mat, "rhosq_year")

# ---- residue submodel draws (optional but handy)
res_base_vec     <- get_param_vec(draws_mat, "res_base")
b_res_stir_vec   <- get_param_vec(draws_mat, "b_res_stir")
gamma_Cr_res_mat <- get_param_mat(draws_mat, "gamma_Cr_res")     # draws × Cr_n
sigma_Cr_res_vec <- get_param_vec(draws_mat, "sigma_Cr_res")
sigma_res_obs_vec <- get_param_vec(draws_mat, "sigma_res_obs")

# ---- Assemble posterior list
post <- list(
  alpha         = alpha_mat,
  beta_stir     = beta_stir_mat,
  beta_cin      = beta_cin_mat,
  beta_dup      = beta_dup_mat,
  beta_res_C    = beta_res_C_mat,

  gamma_B       = gamma_B_arr,
  gamma_S       = gamma_S_arr,
  gamma_F       = gamma_F_arr,

  gamma_Cr      = gamma_Cr_arr,       # draws × A_n × Cr_n
  gamma_Cr_V    = gamma_Cr_V_mat,     # draws × Cr_n

  beta_vol      = beta_vol_vec,
  beta_irr      = beta_irr_mat,
  a_V           = a_V_vec,
  b_V           = b_V_vec,
  sigma_V       = sigma_V_vec,
  sigma_VOL_obs = sigma_VOL_obs_vec,
  beta_res_V    = beta_res_V_vec,

  # C model SDs
  sigma_analyte = sigma_analyte_vec,   # GLOBAL (scalar draws)
  sigma_C_obs   = sigma_C_obs_mat,     # draws × A_n
  mu_log_sigma_C_obs  = mu_log_sigma_C_obs_vec,
  tau_log_sigma_C_obs = tau_log_sigma_C_obs_vec,
  z_log_sigma_C_obs   = z_log_sigma_C_obs_mat,

  # crop SDs
  sigma_Cr      = sigma_Cr_mat,        # draws × A_n
  sigma_Cr_V    = sigma_Cr_V_vec,      # draws

  # GP
  F_year        = F_array,
  etasq_year    = etasq_year_vec,
  rhosq_year    = rhosq_year_vec,

  # residue submodel
  res_base      = res_base_vec,
  b_res_stir    = b_res_stir_vec,
  gamma_Cr_res  = gamma_Cr_res_mat,
  sigma_Cr_res  = sigma_Cr_res_vec,
  sigma_res_obs = sigma_res_obs_vec
)

## =====================================================================
## Summaries
## =====================================================================

# --- STIR (analyte-specific)
alpha_hat      <- colMeans(post$alpha)
beta_stir_mean <- colMeans(post$beta_stir)
beta_stir_hpdi <- apply(post$beta_stir, 2, HPDI, prob = 0.89)
beta_stir_pi   <- apply(post$beta_stir, 2, PI,   prob = 0.95)

# --- Residue on concentration (analyte-specific slope)
beta_resC_coef_mean <- colMeans(post$beta_res_C)
beta_resC_coef_hpdi <- apply(post$beta_res_C, 2, HPDI, prob = 0.89)
beta_resC_coef_pi   <- apply(post$beta_res_C, 2, PI,   prob = 0.95)

beta_resC_eff_mean  <- if (res_is_placeholder) rep(0, A_n) else beta_resC_coef_mean
beta_resC_eff_hpdi  <- if (res_is_placeholder) rbind(rep(0, A_n), rep(0, A_n)) else beta_resC_coef_hpdi
beta_resC_eff_pi    <- if (res_is_placeholder) rbind(rep(0, A_n), rep(0, A_n)) else beta_resC_coef_pi

# --- Analyte effects table
analyte_effects <- analyte_lookup %>%
  mutate(
    alpha_hat        = alpha_hat,

    beta_stir_mean   = beta_stir_mean,
    beta_stir_median = apply(post$beta_stir, 2, median),
    beta_hpdi_low    = beta_stir_hpdi[1, ],
    beta_hpdi_high   = beta_stir_hpdi[2, ],
    beta_pi_low      = beta_stir_pi[1, ],
    beta_pi_high     = beta_stir_pi[2, ],

    beta_resC_mean      = beta_resC_eff_mean,
    beta_resC_median    = if (res_is_placeholder) rep(0, A_n) else apply(post$beta_res_C, 2, median),
    beta_resC_hpdi_low  = beta_resC_eff_hpdi[1, ],
    beta_resC_hpdi_high = beta_resC_eff_hpdi[2, ],
    beta_resC_pi_low    = beta_resC_eff_pi[1, ],
    beta_resC_pi_high   = beta_resC_eff_pi[2, ],

    residue_placeholder = res_is_placeholder
  )

analyte_effects

# --- Crop effects on concentration: one row per (analyte, crop)
crop_effects_conc <- tidyr::expand_grid(
  analyte_lookup %>% select(A, analyte_abbr),
  crop_lookup %>% select(Cr, Crop)
) %>%
  arrange(A, Cr) %>%
  mutate(
    crop_eff_mean   = NA_real_,
    crop_eff_median = NA_real_,
    crop_low89      = NA_real_,
    crop_high89     = NA_real_
  )

for (ii in seq_len(nrow(crop_effects_conc))) {
  a  <- crop_effects_conc$A[ii]
  cr <- crop_effects_conc$Cr[ii]
  x  <- post$gamma_Cr[, a, cr]
  h  <- HPDI(x, prob = 0.89)
  crop_effects_conc$crop_eff_mean[ii]   <- mean(x)
  crop_effects_conc$crop_eff_median[ii] <- median(x)
  crop_effects_conc$crop_low89[ii]      <- h[1]
  crop_effects_conc$crop_high89[ii]     <- h[2]
}

crop_effects_conc

# --- Crop effects on volume: one row per crop
crop_effects_vol <- crop_lookup %>%
  mutate(
    crop_eff_mean   = colMeans(post$gamma_Cr_V),
    crop_eff_median = apply(post$gamma_Cr_V, 2, median),
    crop_low89      = apply(post$gamma_Cr_V, 2, function(x) HPDI(x, prob = 0.89)[1]),
    crop_high89     = apply(post$gamma_Cr_V, 2, function(x) HPDI(x, prob = 0.89)[2])
  )

crop_effects_vol

# --- Residue effect on volume (scalar slope, supported effect is 0 if placeholder)
beta_resV_coef_hpdi <- HPDI(post$beta_res_V, prob = 0.89)
beta_resV_coef_pi   <- PI(post$beta_res_V,   prob = 0.95)

residue_volume_effect <- tibble::tibble(
  beta_resV_mean      = if (res_is_placeholder) 0 else mean(post$beta_res_V),
  beta_resV_median    = if (res_is_placeholder) 0 else median(post$beta_res_V),
  beta_resV_hpdi_low  = if (res_is_placeholder) 0 else beta_resV_coef_hpdi[1],
  beta_resV_hpdi_high = if (res_is_placeholder) 0 else beta_resV_coef_hpdi[2],
  beta_resV_pi_low    = if (res_is_placeholder) 0 else beta_resV_coef_pi[1],
  beta_resV_pi_high   = if (res_is_placeholder) 0 else beta_resV_coef_pi[2],
  residue_placeholder = res_is_placeholder
)

residue_volume_effect

## =====================================================================
## NEW (v1p7p5): convenient sigma checks you can print immediately
## =====================================================================

# 1) Global concentration process SD (latent C_true around mu_C)
sigma_process_C <- tibble::tibble(
  sigma_analyte_mean   = mean(post$sigma_analyte),
  sigma_analyte_median = median(post$sigma_analyte),
  sigma_analyte_low89  = HPDI(post$sigma_analyte, prob = 0.89)[1],
  sigma_analyte_high89 = HPDI(post$sigma_analyte, prob = 0.89)[2]
)

sigma_process_C

# 2) Analyte-specific observation SDs for measured C (sigma_C_obs[A])
sigma_obs_C <- analyte_lookup %>%
  mutate(
    sigma_C_obs_mean   = colMeans(post$sigma_C_obs),
    sigma_C_obs_median = apply(post$sigma_C_obs, 2, median),
    sigma_C_obs_low89  = apply(post$sigma_C_obs, 2, function(x) HPDI(x, prob = 0.89)[1]),
    sigma_C_obs_high89 = apply(post$sigma_C_obs, 2, function(x) HPDI(x, prob = 0.89)[2])
  )

sigma_obs_C

# 3) Hyperparameters behind sigma_C_obs (log scale)
sigma_obs_C_hypers <- tibble::tibble(
  mu_log_sigma_C_obs_mean   = mean(post$mu_log_sigma_C_obs),
  mu_log_sigma_C_obs_median = median(post$mu_log_sigma_C_obs),
  mu_log_sigma_C_obs_low89  = HPDI(post$mu_log_sigma_C_obs, prob = 0.89)[1],
  mu_log_sigma_C_obs_high89 = HPDI(post$mu_log_sigma_C_obs, prob = 0.89)[2],

  tau_log_sigma_C_obs_mean   = mean(post$tau_log_sigma_C_obs),
  tau_log_sigma_C_obs_median = median(post$tau_log_sigma_C_obs),
  tau_log_sigma_C_obs_low89  = HPDI(post$tau_log_sigma_C_obs, prob = 0.89)[1],
  tau_log_sigma_C_obs_high89 = HPDI(post$tau_log_sigma_C_obs, prob = 0.89)[2]
)

sigma_obs_C_hypers


```


## 5a. Posterior effects of residue and crop

The v1p7p5 model includes (i) an analyte-specific residue slope on concentration (`beta_res_C[A]`), (ii) a scalar residue slope on volume (`beta_res_V`), (iii) analyte-specific crop intercepts on concentration (`gamma_Cr[A,Cr]`), and (iv) crop intercepts on volume (`gamma_Cr_V[Cr]`). The plots below provide quick, publication-ready summaries of these terms.

```{r residue effects on concentration}
# Residue effect on concentration by analyte
# ---- Prepare posterior draws ----
post_resC_df <- as.data.frame(post$beta_res_C)
colnames(post_resC_df) <- as.character(analyte_effects$analyte_abbr)

post_resC_long <- post_resC_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "beta_resC"
  )

# ---- Reorder analytes by posterior median (base R, no forcats) ----
analyte_order <- post_resC_long %>%
  group_by(analyte_abbr) %>%
  summarise(med = median(beta_resC)) %>%
  arrange(med) %>%
  pull(analyte_abbr)

post_resC_long$analyte_abbr <- factor(
  post_resC_long$analyte_abbr,
  levels = analyte_order
)

# ---- Ridgeline plot ----
p_resC_ridge <- ggplot(post_resC_long, aes(x = beta_resC, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth=1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Residue effect on concentration",
    y = "Analyte",
    #title = "Posterior residue effects on concentration (v1p7p5)",
    subtitle = if (res_is_placeholder)
      "Residue is currently a placeholder (all zeros), effects are not identifiable"
      else NULL
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_resC_ridge

ggsave(
  filename = "../figs/residue_effect_on_concentration_by_analyte_v1p7p5.jpg",
  plot = p_resC_ridge,
  width = 10,
  height = 7,
  dpi = 300
)

```

```{r residue effect on volume}
# Residue effect on volume (scalar slope) posterior density
# If placeholder, report and plot a spike at zero.

b_resV <- if (isTRUE(res_is_placeholder)) rep(0, 200) else post$beta_res_V

dens <- stats::density(b_resV)

plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "Residue effect on Volume",
  ylab = "Density",
  main = "Posterior residue effect on volume",
  bty  = "l",
  # ensure xlim always shows zero
  xlim = range(c(dens$x, 0))
)
abline(v = 0, lty = 2)

jpeg("../figs/residue_effect_on_volume_density_v1p7p5.jpeg", width = 2000, height = 1300, res = 200)
plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "Residue effect on Volume",
  ylab = "Density",
  main = "Posterior residue effect on volume",
  bty  = "l",
  # ensure xlim always shows zero
  xlim = range(c(dens$x, 0))
)
abline(v = 0, lty = 2)
dev.off()
```

```{r crop effects on concentration}
# ------------------------------------------------------------
# Crop effects on concentration: posterior ridgelines by crop
# post$gamma_Cr is draws × A_n × Cr_n
# ------------------------------------------------------------

# Build a long draw table: one row per (draw, analyte, crop)
crop_levels <- as.character(crop_lookup$Crop)
analyte_levels <- as.character(analyte_lookup$analyte_abbr)

cropC_long <- bind_rows(lapply(seq_len(Cr_n), function(cr) {
  m <- post$gamma_Cr[, , cr, drop = FALSE]      # draws × A × 1
  m2 <- m[, , 1]                               # draws × A
  df <- as.data.frame(m2)
  colnames(df) <- analyte_levels
  df %>%
    pivot_longer(
      cols = everything(),
      names_to = "analyte_abbr",
      values_to = "crop_eff"
    ) %>%
    mutate(Crop = crop_levels[cr])
}))

# Make Crop a factor with your desired ordering
cropC_long$Crop <- factor(cropC_long$Crop, levels = crop_levels)

# Choose ONE analyte order used in every facet (recommended for cross-panel comparison):
analyte_order_global <- cropC_long %>%
  group_by(analyte_abbr) %>%
  summarise(med = median(crop_eff), .groups = "drop") %>%
  arrange(med) %>%
  pull(analyte_abbr)

cropC_long$analyte_abbr <- factor(cropC_long$analyte_abbr, levels = analyte_order_global)

# ------------------------------------------------------------
# NEW: pick a tighter x-axis range from posterior quantiles
# - uses pooled draws across all analytes/crops
# - keeps 0 inside the view
# ------------------------------------------------------------
q_lo <- unname(quantile(cropC_long$crop_eff, probs = 0.005, na.rm = TRUE))
q_hi <- unname(quantile(cropC_long$crop_eff, probs = 0.995, na.rm = TRUE))

# symmetric range around 0, with a small pad so ridges aren't clipped
half_range <- max(abs(c(q_lo, q_hi)), 0.05)   # 0.05 is a safety floor
pad <- 0.05 * half_range
x_lim <- c(-half_range - pad, half_range + pad)

# ---- Faceted ridgeline plot ----
p_cropC_ridge <- ggplot(cropC_long, aes(x = crop_eff, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.8) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.05,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 0.9,
    bandwidth = 0.02
  ) +
  facet_wrap(~ Crop, ncol = 3, scales = "fixed") +
  coord_cartesian(xlim = x_lim) +
  labs(
    x = "Crop effect on concentration",
    y = "Analyte",
    subtitle = "Posterior distributions of analyte-specific crop intercepts"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 12),
    strip.text = element_text(size = 12),
    axis.text.y = element_text(size = 9)
  )

p_cropC_ridge

ggsave(
  filename = "../figs/crop_effects_on_concentration_by_analyte_v1p7p5.jpg",
  plot = p_cropC_ridge,
  width = 9,
  height = 12,
  dpi = 300
)

```

```{r crop effects on volume}
# ------------------------------------------------------------
# Crop effects on volume: ridgeline posterior by crop
# post$gamma_Cr_V is draws × Cr_n
# ------------------------------------------------------------

# Long posterior draw table
post_cropV_df <- as.data.frame(post$gamma_Cr_V)
colnames(post_cropV_df) <- as.character(crop_lookup$Crop)

post_cropV_long <- post_cropV_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "Crop",
    values_to = "crop_eff_V"
  )

# Reorder crops by posterior median (base R, no forcats)
crop_order <- post_cropV_long %>%
  group_by(Crop) %>%
  summarise(med = median(crop_eff_V), .groups = "drop") %>%
  arrange(med) %>%
  pull(Crop)

post_cropV_long$Crop <- factor(post_cropV_long$Crop, levels = crop_order)

# Ridgeline plot
p_cropV_ridge <- ggplot(post_cropV_long, aes(x = crop_eff_V, y = Crop)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Crop effect on volume",
    y = "Crop",
    #title = "Posterior crop effects on volume"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_cropV_ridge

ggsave(
  filename = "../figs/crop_effects_on_volume_v1p7p5.jpg",
  plot = p_cropV_ridge,
  width = 10,
  height = 6,
  dpi = 300
)

```



---



# 4.5 Plot posterior slopes for irrigation-count effects on concentration

```{r irrigation effects on conc}
# ------------------------------------------------------------
# Irrigation-count effects on concentration by analyte
# beta_irr_mat is draws × A_n (MVN prior across analytes in v1p7p5)
# Note: Stan standardizes IRR internally (IRR_z), so beta_irr is per +1 SD(IRR)
# ------------------------------------------------------------

# Long posterior draw table
post_irr_df <- as.data.frame(beta_irr_mat)
colnames(post_irr_df) <- as.character(analytes_keep)

post_irr_long <- post_irr_df %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "Analyte",
    values_to = "beta_irr"
  )

# Reorder analytes by posterior median (base R, no forcats)
analyte_order <- post_irr_long %>%
  dplyr::group_by(Analyte) %>%
  dplyr::summarise(med = median(beta_irr), .groups = "drop") %>%
  dplyr::arrange(med) %>%
  dplyr::pull(Analyte)

post_irr_long$Analyte <- factor(post_irr_long$Analyte, levels = analyte_order)

# Ridgeline plot (match crop style)
p_irr_ridge <- ggplot(post_irr_long, aes(x = beta_irr, y = Analyte)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  ggridges::stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Effect of Irrigation Number on Concentration",
    y = "Analyte"
    # title = "Posterior irrigation-number effects on concentration"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_irr_ridge

ggsave(
  filename = "../figs/irr_effects_on_concentration_v1p7p5.jpg",
  plot = p_irr_ridge,
  width = 10,
  height = 6,
  dpi = 300
)

```

# 5. Plot posterior slopes for STIR effects on concentration

```{r stir effects on conc}
# ------------------------------------------------------------
# STIR effects on concentration by analyte: posterior ridgelines
# Uses post$beta_stir (draws × A_n)
# Overwrites: ../figs/load1p7p5_STIReffects.jpeg
# ------------------------------------------------------------

# ---- Prepare posterior draws ----
post_stir_df <- as.data.frame(post$beta_stir)
colnames(post_stir_df) <- as.character(analyte_effects$analyte_abbr)

post_stir_long <- post_stir_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "beta_stir"
  )

# ---- Order analytes by posterior mean (base R, no forcats) ----
analyte_order <- post_stir_long %>%
  group_by(analyte_abbr) %>%
  summarise(mu = mean(beta_stir), .groups = "drop") %>%
  arrange(mu) %>%
  pull(analyte_abbr)

post_stir_long$analyte_abbr <- factor(
  post_stir_long$analyte_abbr,
  levels = analyte_order
)

# ---- Ridgeline plot ----
p_stir_ridge <- ggplot(post_stir_long, aes(x = beta_stir, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.01
  ) +
  labs(
    x = "Effect of STIR on Concentration",
    y = "Analyte (ordered by mean effect)",
    title = "Analyte-specific STIR effects with posterior medians"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14),
    axis.text.y = element_text(size = 10)
  )

p_stir_ridge

# ---- Save (overwrite existing) ----
ggsave(
  filename = "../figs/load1p7p5_STIReffects.jpeg",
  plot = p_stir_ridge,
  width = 9,
  height = 9,
  dpi = 300
)

```

# ------------------------------------------------------------
# STIR effects on volume
# ------------------------------------------------------------

```{r stir effects on volume}
plot_stir_volume_effect <- function(fit, param = "b_V",
                                    file = NULL,
                                    main = "Posterior effect of STIR on runoff volume",
                                    xlab = "Effect of STIR on runoff volume",
                                    ylab = "Density") {

  if (inherits(fit, "CmdStanMCMC")) {
    draws <- fit$draws(variables = param, format = "matrix")
    if (!(param %in% colnames(draws))) {
      stop("Parameter '", param, "' not found in cmdstan draws.")
    }
    b <- as.numeric(draws[, param])
  } else if (inherits(fit, "stanfit")) {
    s <- rstan::extract(fit, pars = param, permuted = TRUE)
    if (!(param %in% names(s))) stop("Parameter '", param, "' not found in stanfit.")
    b <- as.numeric(s[[param]])
  } else if (is.numeric(fit)) {
    b <- as.numeric(fit)
  } else {
    stop("Unsupported object type for 'fit'. Use a CmdStanMCMC, stanfit, or numeric vector.")
  }

  b <- b[is.finite(b)]

  if (!is.null(file)) {
    jpeg(file, width = 2000, height = 1300, res = 200)
    on.exit(dev.off(), add = TRUE)
  }

  dens <- stats::density(b)

  plot(
    dens,
    type = "l",
    lwd  = 4,
    col  = "darkred",
    xlab = xlab,
    ylab = ylab,
    main = main,
    bty  = "l",
    xlim = range(c(dens$x, 0))
  )

  abline(v = 0, lty = 2)

  invisible(b)
}

# Plot to device
plot_stir_volume_effect(fit, param = "b_V")

# Save to file (v1p7p5 filename)
plot_stir_volume_effect(
  fit,
  param = "b_V",
  file = "../figs/v1p7p5_post_STIR_effect_on_volume.jpg"
)
```


## 6. Generative STIR → Load curves by analyte --------------------------------

```{r load-generative-curves}
## Scenario-based STIR -> Load link + posterior predictive curves ----------
## v1p7p5 (dual ribbons):
##   1) "mean" ribbon = uncertainty in expected mean relationship (link-style)
##   2) "pp" ribbon   = posterior predictive including latent process + obs noise (sim-style)
##
## Notes:
## - Uses sigma_V + sigma_VOL_obs for volume (latent + obs)
## - Uses sigma_analyte + sigma_C_obs[A] for concentration (latent + obs)
## - Residue submodel is NOT used; RES is treated as an input covariate.
## - v1p7p5: irrigation effect is analyte-specific and IRR is standardized (IRR_z).

# 1. Back-transform stats (only computed once)
c_stats <- d_mod %>%
  dplyr::group_by(A) %>%
  dplyr::summarize(
    c_mean = mean(Result_mg_L, na.rm = TRUE),
    c_sd   = sd(Result_mg_L,   na.rm = TRUE),
    .groups = "drop"
  )

V_mean <- mean(d_mod$Volume, na.rm = TRUE)
V_sd   <- sd(d_mod$Volume, na.rm = TRUE)

# STIR raw mean/sd for Season_STIR_toDate
stir_mean <- mean(d_mod$Season_STIR_toDate, na.rm = TRUE)
stir_sd   <- sd(d_mod$Season_STIR_toDate, na.rm = TRUE)

# Empirical means for averaging when left NULL
IRR_bar_raw <- mean(d_mod$irr_num, na.rm = TRUE)
DUP_bar     <- mean(d_mod$dup,     na.rm = TRUE)

# v1p7p5: IRR standardization to match Stan IRR_z
irr_mean <- mean(d_mod$irr_num, na.rm = TRUE)
irr_sd   <- sd(d_mod$irr_num,   na.rm = TRUE)
if (!is.finite(irr_sd) || irr_sd == 0) irr_sd <- 1
IRR_bar_z <- (IRR_bar_raw - irr_mean) / irr_sd

# Empirical mean residue proportion when res_z is NULL
RES_bar <- if ("RES_input" %in% names(d_mod)) {
  mean(d_mod$RES_input, na.rm = TRUE)
} else if ("RES" %in% names(d_mod)) {
  mean(d_mod$RES, na.rm = TRUE)
} else if ("residue_prop" %in% names(d_mod)) {
  mean(d_mod$residue_prop, na.rm = TRUE)
} else if ("RES_z" %in% names(d_mod)) {
  mean(d_mod$RES_z, na.rm = TRUE)
} else {
  NA_real_
}

if (!is.finite(RES_bar)) stop("Could not compute RES_bar: d_mod needs RES_input, RES, residue_prop, or RES_z as a residue proportion in (0,1).")
if (RES_bar <= 0 || RES_bar >= 1) stop("RES_bar is not in (0,1). Check that residue is stored as a proportion, not percent.")

# Crop weights for "average over crops" scenario
crop_wts <- d_mod %>%
  dplyr::count(Cr, name = "n") %>%
  dplyr::mutate(w = n / sum(n)) %>%
  dplyr::arrange(Cr)

simulate_load_curve_single <- function(
  post,
  analyte,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block      = NULL,
  sampler    = NULL,
  flume      = NULL,
  irrigation = NULL,
  duplicate  = NULL,
  crop       = NULL,
  res_z      = NULL,   # NULL -> RES_bar; 0 -> 1e-4; (0,1) fixed proportion
  cin_z      = 0,
  year       = NULL,
  n_draws    = 2000,
  ci_prob    = 0.95
) {

  if (!is.numeric(ci_prob) || length(ci_prob) != 1L || ci_prob <= 0 || ci_prob >= 1) {
    stop("ci_prob must be a single numeric value strictly between 0 and 1.")
  }

  # ---- required sigmas in post (fail fast if missing) ----
  need <- c("sigma_V", "sigma_VOL_obs", "sigma_analyte")
  miss <- setdiff(need, names(post))
  if (length(miss) > 0) stop("post is missing required parameter(s): ", paste(miss, collapse = ", "))

  # concentration obs sd is analyte-specific
  if (!("sigma_C_obs" %in% names(post))) {
    stop("post is missing sigma_C_obs (analyte-specific concentration observation SD).")
  }

  # Map analyte label -> index A
  j_idx <- analyte_lookup$A[analyte_lookup$analyte_abbr == analyte]
  if (length(j_idx) != 1L) stop("analyte not found in analyte_lookup")

  # Stats for this analyte
  stats_j  <- dplyr::filter(c_stats, A == j_idx)
  c_mean_j <- stats_j$c_mean
  c_sd_j   <- stats_j$c_sd

  # STIR in z units
  STIR_z_seq <- (STIR_raw_seq - stir_mean) / stir_sd

  # Total draws and subsample index
  n_draws_total <- nrow(post$alpha)
  draw_idx <- if (n_draws_total > n_draws) sample.int(n_draws_total, n_draws) else seq_len(n_draws_total)
  n_s <- length(draw_idx)

  # ---------------- Random effects / covariate settings ----------------

  # Block (Rep)
  if (is.null(block)) {
    gamma_B_draws <- post$gamma_B[draw_idx, j_idx, , drop = FALSE]   # draws × 1 × B_n
    gamma_B_star  <- apply(gamma_B_draws, 1, mean)
    block_label   <- "avg_all_blocks"
  } else {
    B_idx <- block_lookup$B[block_lookup$Rep == block]
    if (length(B_idx) != 1L) stop("block not found in block_lookup")
    gamma_B_star <- post$gamma_B[draw_idx, j_idx, B_idx]
    block_label  <- as.character(block)
  }

  # Sampler method
  if (is.null(sampler)) {
    gamma_S_draws <- post$gamma_S[draw_idx, j_idx, , drop = FALSE]
    gamma_S_star  <- apply(gamma_S_draws, 1, mean)
    sampler_label <- "avg_all_samplers"
  } else {
    S_idx <- sampler_lookup$S[sampler_lookup$SampleMethod == sampler]
    if (length(S_idx) != 1L) stop("sampler not found in sampler_lookup")
    gamma_S_star  <- post$gamma_S[draw_idx, j_idx, S_idx]
    sampler_label <- as.character(sampler)
  }

  # Flume type
  if (is.null(flume)) {
    gamma_F_draws <- post$gamma_F[draw_idx, j_idx, , drop = FALSE]
    gamma_F_star  <- apply(gamma_F_draws, 1, mean)
    flume_label   <- "avg_all_flumes"
  } else {
    Fu_idx <- flume_lookup$Fu[flume_lookup$FlumeMethod == flume]
    if (length(Fu_idx) != 1L) stop("flume not found in flume_lookup")
    gamma_F_star  <- post$gamma_F[draw_idx, j_idx, Fu_idx]
    flume_label   <- as.character(flume)
  }

  # Irrigation (numeric)
  # IMPORTANT: convert raw irrigation count -> IRR_z scale that Stan uses
  if (is.null(irrigation)) {
    IRR_star_raw     <- IRR_bar_raw
    IRR_star_z       <- IRR_bar_z
    irrigation_label <- "avg_all_irrigations"
  } else {
    irr_val <- irrigation_lookup$irr_num[irrigation_lookup$Irrigation == irrigation]
    if (length(irr_val) != 1L) stop("irrigation not found in irrigation_lookup")
    IRR_star_raw     <- irr_val
    IRR_star_z       <- (IRR_star_raw - irr_mean) / irr_sd
    irrigation_label <- as.character(irrigation)
  }

  # Duplicate status (numeric 0/1 or mean)
  if (is.null(duplicate)) {
    DUP_star        <- DUP_bar
    duplicate_label <- "avg_duplicates"
  } else {
    DUP_star        <- as.numeric(duplicate)
    duplicate_label <- ifelse(as.logical(duplicate), "TRUE", "FALSE")
  }

  # Crop: affects BOTH volume and concentration
  if (is.null(crop)) {
    w <- crop_wts$w
    if (length(w) != ncol(post$gamma_Cr_V)) stop("crop weight length != Cr_n for gamma_Cr_V")
    if (length(w) != dim(post$gamma_Cr)[3]) stop("crop weight length != Cr_n for gamma_Cr")

    cropV_draws <- post$gamma_Cr_V[draw_idx, , drop = FALSE]
    gamma_CrV_star <- as.numeric(cropV_draws %*% w)

    cropC_draws <- post$gamma_Cr[draw_idx, j_idx, , drop = FALSE]  # draws × 1 × Cr_n
    cropC_mat   <- matrix(cropC_draws, nrow = n_s, ncol = length(w))
    gamma_CrC_star <- as.numeric(cropC_mat %*% w)

    crop_label <- "avg_over_crops_weighted"
  } else {
    Cr_idx <- crop_lookup$Cr[crop_lookup$Crop == crop]
    if (length(Cr_idx) != 1L) stop("crop not found in crop_lookup (check exact spelling/levels)")
    gamma_CrV_star <- post$gamma_Cr_V[draw_idx, Cr_idx]
    gamma_CrC_star <- post$gamma_Cr[draw_idx, j_idx, Cr_idx]
    crop_label <- as.character(crop)
  }

  # ---------------- Residue handling ----------------
  if (!is.null(res_z) && (length(res_z) != 1L || !is.numeric(res_z))) {
    stop("res_z must be NULL or a single numeric value.")
  }

  if (is.null(res_z)) {
    res_prop <- RES_bar
    res_mode <- "avg_observed"
  } else {
    res_prop <- as.numeric(res_z)
    if (!is.finite(res_prop)) stop("res_z must be finite when not NULL.")
    if (identical(res_prop, 0)) res_prop <- 1e-4
    res_mode <- "fixed"
  }
  if (res_prop <= 0 || res_prop >= 1) stop("Residue proportion must be strictly between 0 and 1. Got: ", res_prop)

  # ---------------- Posterior pieces for this analyte ----------------
  alpha_j      <- post$alpha[draw_idx, j_idx]
  beta_stirj   <- post$beta_stir[draw_idx, j_idx]
  beta_cinj    <- post$beta_cin[draw_idx, j_idx]
  beta_dupj    <- post$beta_dup[draw_idx, j_idx]
  beta_resC_j  <- post$beta_res_C[draw_idx, j_idx]

  beta_vol     <- post$beta_vol[draw_idx]

  # v1p7p5: analyte-specific irrigation effect
  beta_irrj    <- post$beta_irr[draw_idx, j_idx]

  a_V          <- post$a_V[draw_idx]
  b_V          <- post$b_V[draw_idx]
  beta_res_V   <- post$beta_res_V[draw_idx]

  # sigmas
  sigma_V      <- post$sigma_V[draw_idx]
  sigma_VOLobs <- post$sigma_VOL_obs[draw_idx]
  sigma_Cproc  <- post$sigma_analyte[draw_idx]
  sigma_Cobs_j <- post$sigma_C_obs[draw_idx, j_idx]

  # GP on year
  if (is.null(year)) {
    F_sub  <- post$F_year[draw_idx, , j_idx, drop = FALSE]  # draws × Y × 1
    k_star <- apply(F_sub, 1, mean)
    year_label <- "avg_all_years"
  } else {
    Y_idx <- match(year, years_used)
    if (is.na(Y_idx)) stop("requested year not in years_used")
    k_star <- post$F_year[draw_idx, Y_idx, j_idx]
    year_label <- as.character(year)
  }

  # ---------------- Loop over STIR values ----------------
  n_x <- length(STIR_z_seq)

  # link-style (mean function) summaries
  load_mean <- numeric(n_x)
  load_low  <- numeric(n_x)
  load_high <- numeric(n_x)

  # posterior predictive summaries (latent + obs noise)
  load_pp_mean <- numeric(n_x)
  load_pp_low  <- numeric(n_x)
  load_pp_high <- numeric(n_x)

  for (k in seq_len(n_x)) {
    stz <- STIR_z_seq[k]
    RES_star <- rep(res_prop, n_s)

    # mean functions on z-scale
    muV_z <- a_V +
      b_V * stz +
      beta_res_V * RES_star +
      gamma_CrV_star

    muC_z <- alpha_j +
      beta_stirj * stz +
      beta_cinj  * cin_z +
      beta_vol   * muV_z +
      beta_irrj  * IRR_star_z +     # << v1p7p5 IRR_z scale
      beta_dupj  * DUP_star +
      beta_resC_j * RES_star +
      gamma_CrC_star +
      gamma_B_star +
      gamma_S_star +
      gamma_F_star +
      k_star

    # ---- (1) Link-style: mean relationship only ----
    C_raw_mu <- c_mean_j + c_sd_j * muC_z
    V_raw_mu <- V_mean   + V_sd   * muV_z
    L_link   <- C_raw_mu * V_raw_mu / 1000

    load_mean[k] <- mean(L_link)
    hpdi_k       <- HPDI(L_link, prob = ci_prob)
    load_low[k]  <- hpdi_k[1]
    load_high[k] <- hpdi_k[2]

    # ---- (2) Posterior predictive: latent + observation noise ----
    V_true_z <- rnorm(n_s, mean = muV_z, sd = sigma_V)
    V_obs_z  <- rnorm(n_s, mean = V_true_z, sd = sigma_VOLobs)

    C_true_z <- rnorm(n_s, mean = muC_z, sd = sigma_Cproc)
    C_obs_z  <- rnorm(n_s, mean = C_true_z, sd = sigma_Cobs_j)

    C_raw_pp <- c_mean_j + c_sd_j * C_obs_z
    V_raw_pp <- V_mean   + V_sd   * V_obs_z
    L_pp     <- C_raw_pp * V_raw_pp / 1000

    load_pp_mean[k] <- mean(L_pp)
    hpdi_pp         <- HPDI(L_pp, prob = ci_prob)
    load_pp_low[k]  <- hpdi_pp[1]
    load_pp_high[k] <- hpdi_pp[2]
  }

  data.frame(
    analyte            = analyte,
    STIR_raw           = STIR_raw_seq,

    # mean relationship (link-style)
    load_mean          = load_mean,
    load_low           = load_low,
    load_high          = load_high,

    # posterior predictive (sim-style)
    load_pp_mean       = load_pp_mean,
    load_pp_low        = load_pp_low,
    load_pp_high       = load_pp_high,

    block_level        = block_label,
    sampler_level      = sampler_label,
    flume_level        = flume_label,
    irrigation_level   = irrigation_label,
    duplicate_level    = duplicate_label,
    crop_level         = crop_label,
    residue_mode       = res_mode,
    res_z              = if (is.null(res_z)) NA_real_ else as.numeric(res_z),
    res_fixed_prop     = res_prop,
    cin_z              = cin_z,
    year_level         = year_label,
    ci_prob            = ci_prob
  )
}

simulate_load_curves <- function(
  post,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = NULL,
  crop         = NULL,
  res_z        = NULL,
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
) {
  if (is.null(analytes)) analytes <- as.character(analyte_lookup$analyte_abbr)

  res_list <- lapply(analytes, function(an) {
    simulate_load_curve_single(
      post         = post,
      analyte      = an,
      STIR_raw_seq = STIR_raw_seq,
      block        = block,
      sampler      = sampler,
      flume        = flume,
      irrigation   = irrigation,
      duplicate    = duplicate,
      crop         = crop,
      res_z        = res_z,
      cin_z        = cin_z,
      year         = year,
      n_draws      = n_draws,
      ci_prob      = ci_prob
    )
  })

  dplyr::bind_rows(res_list)
}

compute_observed_load_points <- function(
  d_mod,
  analytes,
  block = NULL,
  sampler = NULL,
  flume = NULL,
  irrigation = NULL,
  duplicate = NULL,
  crop = NULL,
  year = NULL
) {

  df <- d_mod

  # Ensure analyte label exists
  if (!("analyte_abbr" %in% names(df))) {
    if ("A" %in% names(df)) {
      df$analyte_abbr <- analyte_lookup$analyte_abbr[match(df$A, analyte_lookup$A)]
    } else {
      stop("compute_observed_load_points(): need d_mod$analyte_abbr or d_mod$A.")
    }
  }

  # Filter analytes
  df <- df[df$analyte_abbr %in% analytes, , drop = FALSE]

  # Optional filters (apply only if columns exist)
  if (!is.null(block)      && "Rep" %in% names(df))           df <- df[df$Rep == block, , drop = FALSE]
  if (!is.null(sampler)    && "SampleMethod" %in% names(df))  df <- df[df$SampleMethod == sampler, , drop = FALSE]
  if (!is.null(flume)      && "FlumeMethod" %in% names(df))   df <- df[df$FlumeMethod == flume, , drop = FALSE]
  if (!is.null(irrigation) && "irr_num" %in% names(df))       df <- df[df$irr_num == irrigation, , drop = FALSE]
  if (!is.null(duplicate)  && "dup" %in% names(df))           df <- df[df$dup == as.numeric(duplicate), , drop = FALSE]

  if (!is.null(crop)) {
    if ("Crop" %in% names(df)) df <- df[df$Crop == crop, , drop = FALSE]
    if ("Cr" %in% names(df)) {
      Cr_idx <- crop_lookup$Cr[crop_lookup$Crop == crop][1]
      df <- df[df$Cr == Cr_idx, , drop = FALSE]
    }
  }

  if (!is.null(year)) {
    if ("year" %in% names(df)) df <- df[df$year == year, , drop = FALSE]
    if ("Year" %in% names(df)) df <- df[df$Year == year, , drop = FALSE]
  }

  req <- c("Result_mg_L", "Volume", "Season_STIR_toDate", "analyte_abbr")
  miss <- setdiff(req, names(df))
  if (length(miss) > 0) stop("compute_observed_load_points(): missing: ", paste(miss, collapse = ", "))

  df <- df[is.finite(df$Result_mg_L) & is.finite(df$Volume) & is.finite(df$Season_STIR_toDate), , drop = FALSE]

  df$STIR_raw <- df$Season_STIR_toDate
  df$load_obs <- df$Result_mg_L * df$Volume / 1000  # mg -> g

  # CRITICAL: provide the faceting variable used by predictions
  df$analyte <- as.character(df$analyte_abbr)

  df
}

simulate_load_curves_with_obs <- function(
  post,
  d_mod,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = NULL,
  crop         = NULL,
  res_z        = NULL,
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
) {

  pred <- simulate_load_curves(
    post         = post,
    analytes     = analytes,
    STIR_raw_seq = STIR_raw_seq,
    block        = block,
    sampler      = sampler,
    flume        = flume,
    irrigation   = irrigation,
    duplicate    = duplicate,
    crop         = crop,
    res_z        = res_z,
    cin_z        = cin_z,
    year         = year,
    n_draws      = n_draws,
    ci_prob      = ci_prob
  )

  obs <- compute_observed_load_points(
    d_mod      = d_mod,
    analytes   = unique(pred$analyte),
    block      = block,
    sampler    = sampler,
    flume      = flume,
    irrigation = irrigation,
    duplicate  = duplicate,
    crop       = crop,
    year       = year
  )

  list(pred = pred, obs = obs)
}
```


## Scenario-based STIR → Load Simulation Functions

### Parameter definitions and behavior

The simulation functions allow you to generate posterior predicted analyte loads across a range of STIR values under different user-defined “scenarios.” Each scenario variable may be explicitly specified or left blank. When a parameter is left blank (NULL), the model automatically averages over that variable by using the posterior mean of its random effect or the empirical mean of the covariate.

**analyte**

* Character analyte abbreviation (for example, `"TP"`, `"NO3"`, `"TDS"`).
* Default behavior: uses *all analytes*.
* Valid levels come from `analyte_lookup$analyte_abbr`:
  NH4, ICP, NO3, NOx, NO2, NPOC, OP, Se, TDS, TKN, TN, TP, TSP, TSS.

**block**

* Field block or replication factor, matching `block_lookup$Rep` (typically `"1"` or `"2"`).
* If left NULL: averaged over blocks via the posterior mean of the random effect `gamma_B`.

**sampler**

* Sampler method type, matching `sampler_lookup$SampleMethod` (GB, GB3, GBH, GBW, ISC, LC).
* If left NULL: averaged over sampler methods via the posterior mean of `gamma_S`.

**flume**

* Flume hardware type, matching `flume_lookup$FlumeMethod`:
  “10 V”, “10V”, “60 V Trap”, “7 V”, “8 V”, “Weir”.
* If left NULL: averaged using the posterior mean of `gamma_F`.

**irrigation**

* Irrigation-event identity, matching `irrigation_lookup$Irrigation`:
  1, 10, 2, 3, 4, 5, 6, 7, 8, 9, S1, S2.
* IRR is a fixed-effect covariate, not a random effect.
* If left NULL: the function uses the empirical mean of `irr_num` in the dataset.
* v1p7p5: internally converted to standardized IRR_z to match the Stan model.

**duplicate**

* Laboratory duplicate indicator (TRUE or FALSE).
* If NULL: uses the empirical mean duplicate rate across the dataset.

**cin_z**

* Standardized inflow concentration for the selected analyte.
* Defaults to 0 (“average inflow concentration”).

**STIR_raw_seq**

* The raw STIR values over which predictions are generated (for example, 0–300).

**n_draws**

* Number of posterior samples used for simulation, for computational efficiency.

**ci_prob**

* Credible interval probability for load estimates (default 0.95).

---


### Output

The functions return a tidy data frame containing:

* analyte
* STIR_raw
* load_mean (posterior mean load in mg)
* load_low (lower HPDI)
* load_high (upper HPDI)
* load_pp_mean (posterior predictive mean load)
* load_pp_low
* load_pp_high

This output is designed for straightforward faceted visualization in `ggplot2`.

```{r load-generative-curves-user-guide}
###############################################################
## USER INPUT GUIDE FOR `simulate_load_curves_with_obs()` (v1p7p5)
###############################################################

# ======================================================================
# Example scenario run (v1p7p5 dual ribbons)
# ======================================================================

res_scenario <- simulate_load_curves_with_obs(
  post         = post,
  d_mod        = d_mod,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = NULL,
  crop         = NULL,
  res_z        = 0.10,   # FIXED residue proportion (0.10 = 10% cover). Use NULL to average observed residue.
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
)

load_pred_scenario <- res_scenario$pred
load_obs_scenario  <- res_scenario$obs

# Build a dynamic subtitle from scenario labels
scenario_info <- load_pred_scenario %>%
  dplyr::distinct(
    year_level,
    block_level, sampler_level, flume_level,
    irrigation_level, duplicate_level,
    crop_level, residue_mode, res_fixed_prop, cin_z
  )

residue_text <- if (scenario_info$residue_mode == "avg_observed") {
  sprintf("avg observed %.1f%%", 100 * scenario_info$res_fixed_prop)
} else {
  sprintf("fixed %.1f%%", 100 * scenario_info$res_fixed_prop)
}

subtitle_text <- sprintf(
  "Year: %s; Block: %s; Sampler: %s;\nFlume: %s; Irrigation: %s;\nDuplicate: %s; Crop: %s;\nResidue: %s; Inflow(z)=%.2f",
  scenario_info$year_level,
  scenario_info$block_level,
  scenario_info$sampler_level,
  scenario_info$flume_level,
  scenario_info$irrigation_level,
  scenario_info$duplicate_level,
  scenario_info$crop_level,
  residue_text,
  scenario_info$cin_z
)

ci_pct <- round(100 * unique(load_pred_scenario$ci_prob)[1])

# Dual-ribbon plot:
#   - outer ribbon (pp) = latent process + observation noise
#   - inner ribbon (mean) = uncertainty in expected relationship
p_load <- ggplot(load_pred_scenario, aes(x = STIR_raw)) +
  geom_ribbon(
    aes(ymin = load_pp_low, ymax = load_pp_high),
    fill = "gray75", alpha = 0.35, linewidth = 0
  ) +
  geom_ribbon(
    aes(ymin = load_low, ymax = load_high),
    fill = "gray45", alpha = 0.35, linewidth = 0
  ) +
  geom_line(aes(y = load_mean), linewidth = 1.1) +
  geom_point(
    data = load_obs_scenario,
    aes(x = STIR_raw, y = load_obs, group = analyte),
    inherit.aes = FALSE,
    size = 2.0, alpha = 0.9
  ) +
  facet_wrap(~ analyte, scales = "free_y", ncol = 4) +
  labs(
    x = "Seasonal STIR",
    y = sprintf("Load (g per event), ribbons show %d%% intervals", ci_pct),
    title = "STIR → Load curves by analyte",
    subtitle = paste0(
      subtitle_text,
      "\nDark ribbon: mean relationship uncertainty. Light ribbon: posterior predictive (process + observation)."
    )
  ) +
  theme_classic(base_size = 18) +
  theme(
    strip.text    = element_text(size = 16, face = "bold"),
    axis.text     = element_text(size = 14),
    axis.title    = element_text(size = 18),
    plot.title    = element_text(size = 22, face = "bold"),
    plot.subtitle = element_text(size = 14),
    panel.spacing = unit(1.2, "lines"),
    plot.margin   = margin(15, 20, 15, 20)
  )

p_load
```


```{r save load fig}
# Save to figs folder (v1p7p5)
ggsave(
  filename = "../figs/load1p7p5_STIR_load_curves.jpeg",
  plot = p_load,
  width = 12, height = 10, dpi = 300
)
```

Plot effects of time (i.e., Year)
```{r year covar plot}
# ------------------------------------------------------------
# GP covariance helper (matches Stan cov_GPL2)
# K(d) = etasq * exp(-rhosq * d^2)
# NOTE: Stan adds a diagonal nugget delta (e.g., 0.01) separately.
# ------------------------------------------------------------
cov_fun <- function(etasq, rhosq, d) {
  etasq <- as.numeric(etasq)
  rhosq <- as.numeric(rhosq)
  d     <- as.numeric(d)
  etasq * exp(-rhosq * d^2)
}

if (!exists("col_alpha")) {
  col_alpha <- function(col, alpha = 1) grDevices::adjustcolor(col, alpha.f = alpha)
}
# ============================================================
# Year-distance covariance functions (prior vs posterior)
# Self-contained (defines x_seq, d_max, y_max, posterior subsample)
# ============================================================

# --- Safety checks ---
stopifnot(exists("years_used"))
stopifnot(exists("post"))

# ---- ensure needed objects exist in THIS chunk ----
d_max <- max(dist(years_used))
x_seq <- seq(0, d_max, length.out = 200)

# Prior/post draws (define etasq_post/rhosq_post if you use them below)
n_prior <- 50
n_post  <- 50

etasq_prior <- rexp(n_prior, rate = 2)
rhosq_prior <- rexp(n_prior, rate = 0.5)

etasq_all <- as.numeric(post$etasq_year)
rhosq_all <- as.numeric(post$rhosq_year)

set.seed(1)
take <- if (length(etasq_all) > n_post) sample.int(length(etasq_all), n_post) else seq_along(etasq_all)
etasq_post <- etasq_all[take]
rhosq_post <- rhosq_all[take]

# Robust posterior-mean curve: always returns length(x_seq)
pm_cov <- vapply(
  x_seq,
  FUN = function(x) mean(cov_fun(etasq_all, rhosq_all, x), na.rm = TRUE),
  FUN.VALUE = numeric(1)
)
pm_mean <- pm_cov

# y_max guard
y_max <- max(
  cov_fun(etasq_prior, rhosq_prior, 0),
  cov_fun(etasq_post,  rhosq_post,  0),
  pm_mean,
  na.rm = TRUE
)

# ---------------------------------
# Save figure
# ---------------------------------
jpeg("../figs/year_covariance_gp_v1p7p5.jpeg", width = 2000, height = 1500, res = 220)

plot(
  NULL,
  xlab = "distance (years)",
  ylab = "covariance",
  xlim = c(0, d_max),
  ylim = c(0, y_max),
  main = "Year GP covariance: prior vs posterior"
)

# PRIOR curves (grey/black)
for (i in seq_len(n_prior)) {
  curve(
    cov_fun(etasq_prior[i], rhosq_prior[i], x),
    from = 0, to = d_max,
    add = TRUE,
    col = col_alpha("black", 0.20),
    lwd = 2
  )
}

# POSTERIOR curves (firebrick)
for (i in seq_len(length(etasq_post))) {
  curve(
    cov_fun(etasq_post[i], rhosq_post[i], x),
    from = 0, to = d_max,
    add = TRUE,
    col = col_alpha("firebrick", 0.35),
    lwd = 2
  )
}

# Posterior mean (dashed dark grey)
lines(x_seq, pm_mean, col = "grey30", lwd = 4, lty = 2)

legend(
  "topright",
  legend = c("Prior draws", "Posterior draws", "Posterior mean"),
  col    = c(col_alpha("black", 0.4), col_alpha("firebrick", 0.5), "grey30"),
  lwd    = c(2, 2, 4),
  lty    = c(1, 1, 2),
  bty    = "n"
)

dev.off()

# ---------------------------------
# Also display inline
# ---------------------------------
plot(
  NULL,
  xlab = "distance (years)",
  ylab = "covariance",
  xlim = c(0, d_max),
  ylim = c(0, y_max),
  main = "Year GP covariance: prior vs posterior"
)

for (i in seq_len(n_prior)) {
  curve(
    cov_fun(etasq_prior[i], rhosq_prior[i], x),
    from = 0, to = d_max,
    add = TRUE,
    col = col_alpha("black", 0.20),
    lwd = 2
  )
}

for (i in seq_len(length(etasq_post))) {
  curve(
    cov_fun(etasq_post[i], rhosq_post[i], x),
    from = 0, to = d_max,
    add = TRUE,
    col = col_alpha("firebrick", 0.35),
    lwd = 2
  )
}

lines(x_seq, pm_mean, col = "grey30", lwd = 4, lty = 2)

legend(
  "topright",
  legend = c("Prior draws", "Posterior draws", "Posterior mean"),
  col    = c(col_alpha("black", 0.4), col_alpha("firebrick", 0.5), "grey30"),
  lwd    = c(2, 2, 4),
  lty    = c(1, 1, 2),
  bty    = "n"
)

```


```{r analyte persistence plot}
## ============================================================
## Analyte persistence: lag-1 correlation of latent year effect
## v1p7p5 (uses post$F_year: draws × Y × A)
## Ridgeline replacement (overwrites same filename pattern)
## ============================================================

F_arr <- post$F_year
stopifnot(length(dim(F_arr)) == 3)

n_draws <- dim(F_arr)[1]
Y_n     <- dim(F_arr)[2]
A_n     <- dim(F_arr)[3]

# compute lag-1 correlations for each analyte and draw
lag1_mat <- matrix(NA_real_, nrow = n_draws, ncol = A_n)

for (a in seq_len(A_n)) {
  Fa <- F_arr[, , a]  # draws × Y
  for (s in seq_len(n_draws)) {
    z <- Fa[s, ]
    if (is.finite(sd(z)) && sd(z) > 0) {
      lag1_mat[s, a] <- stats::cor(z[-Y_n], z[-1])
    }
  }
}

# ---- Build long posterior draw table (no forcats) ----
lag1_df <- as.data.frame(lag1_mat)
colnames(lag1_df) <- as.character(analyte_lookup$analyte_abbr)

lag1_long <- lag1_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "lag1_cor"
  ) %>%
  filter(is.finite(lag1_cor))

# order analytes by posterior mean lag-1 correlation
analyte_order <- lag1_long %>%
  group_by(analyte_abbr) %>%
  summarise(mu = mean(lag1_cor), .groups = "drop") %>%
  arrange(mu) %>%
  pull(analyte_abbr)

lag1_long$analyte_abbr <- factor(lag1_long$analyte_abbr, levels = analyte_order)

# ---- Ridgeline plot ----
p_persist_ridge <- ggplot(lag1_long, aes(x = lag1_cor, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Lag-1 correlation (F_year)",
    y = "Analyte (ordered by mean persistence)",
    title = "Analyte-specific persistence of temporal deviations"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14),
    axis.text.y = element_text(size = 10)
  )

p_persist_ridge

# Save to figs folder
ggsave(
  filename = "../figs/analyte_persistence_precis_v1p7p5.jpeg",
  plot = p_persist_ridge,
  width = 12,
  height = 10,
  dpi = 300
)

```



```{r year effects plot}
## ============================================================
## Year effects plot: F_year by analyte (v1p7p5)
## post$F_year is draws × Y × A
## ============================================================

F_arr <- post$F_year
stopifnot(length(dim(F_arr)) == 3)

Y_n <- dim(F_arr)[2]
stopifnot(length(years_used) == Y_n)

A_n <- dim(F_arr)[3]

# posterior means and intervals for F_year
F_mean <- apply(F_arr, c(2, 3), mean, na.rm = TRUE)             # Y × A
F_hpdi <- apply(F_arr, c(2, 3), HPDI, prob = 0.89)              # 2 × Y × A

# Build long data frame
F_df <- expand.grid(
  Y = seq_len(Y_n),
  A = seq_len(A_n)
) %>%
  dplyr::mutate(
    effect_mean = as.vector(F_mean),          # length Y*A
    hpdi_low    = as.vector(F_hpdi[1, , ]),   # length Y*A
    hpdi_high   = as.vector(F_hpdi[2, , ])    # length Y*A
  ) %>%
  dplyr::left_join(analyte_lookup, by = "A") %>%
  dplyr::mutate(Year = years_used[Y])

gg_F <- ggplot(
  F_df,
  aes(x = Year, y = effect_mean, ymin = hpdi_low, ymax = hpdi_high)
) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_ribbon(alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ analyte_abbr, scales = "free_y", ncol = 4) +
  labs(
    x = "Year",
    y = "Latent temporal deviation (z concentration scale)",
    title = "Year-to-year latent deviations by analyte"
  ) +
  theme_classic(base_size = 14)

gg_F

# Save to figs folder
ggsave(
  filename = "../figs/yearly_latent_deviations_v1p7p5.jpeg",
  plot = gg_F,
  width = 12, height = 10, dpi = 300
)

```


## Annual Load Table and Plot

```{r pred data generation}
# ============================================================
# v1p7p5: Build a prediction frame that includes "missing chemistry years"
# Drop-in replacement with:
#   1) robust irrigation mapping (keeps S1/S2, avoids coercion warnings)
#   2) diagnostics explaining why years/rows get dropped
#   3) optional cin_z fill (default: NA cin_z -> 0)
#   4) fields: Crop + residue placeholder (RES_z)
#      - Crop is required for Stan (Cr index)
#      - Residue is required for Stan (RES predictor)
#        For now we set RES_z = 0 (placeholder; no impact)
#
# IMPORTANT v1p7p5 note:
#   - Irrigation is standardized later as irr_z to match the Stan data scale.
# ============================================================
build_d_pred <- function(
  wq_clean,
  event_id_col = "orig_row",
  keep_outflow_only = TRUE,
  drop_no_runoff = TRUE,
  cin_na_to_zero = TRUE,
  diagnostics = TRUE,
  res_na_to_zero = TRUE  # if RES_z ever exists, fill NA -> 0
) {

  # required columns
  needed <- c(
    "Year","Treatment","analyte_abbr","Season_STIR_toDate",
    "Rep","SampleMethod","FlumeMethod","Irrigation","Duplicate","cin_z",
    "Crop",
    event_id_col
  )
  miss <- setdiff(needed, names(wq_clean))
  if (length(miss) > 0) stop("wq_clean missing columns: ", paste(miss, collapse = ", "))

  d0 <- wq_clean

  if (diagnostics) {
    message("\n---- Year coverage in raw wq_clean (before any filters) ----")
    print(wq_clean %>% dplyr::count(Year) %>% dplyr::arrange(Year))
  }

  # ------------------------------------------------------------
  # Filters (apply BEFORE transmute)
  # ------------------------------------------------------------
  if (keep_outflow_only && "InflowOutflow" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(.data$InflowOutflow == "OUT")
  }

  if (drop_no_runoff && "NoRunoff" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(!isTRUE(.data$NoRunoff))
  }

  if (diagnostics) {
    message("\n---- Year coverage after keep_outflow_only / drop_no_runoff filters ----")
    print(d0 %>% dplyr::count(Year) %>% dplyr::arrange(Year))

    message("\n---- Years removed by filters (raw vs filtered) ----")
    raw_years  <- wq_clean %>% dplyr::distinct(Year)
    kept_years <- d0       %>% dplyr::distinct(Year)
    removed <- dplyr::anti_join(raw_years, kept_years, by = "Year") %>% dplyr::arrange(Year)
    print(removed)
  }

  # ------------------------------------------------------------
  # Robust irrigation mapping (no "NAs introduced by coercion")
  # ------------------------------------------------------------
  irr_levels <- sort(unique(as.character(d0$Irrigation)))

  irr_map <- tibble::tibble(Irrigation_chr = irr_levels) %>%
    dplyr::mutate(
      irr_num = dplyr::case_when(
        Irrigation_chr == "S1" ~ 11,
        Irrigation_chr == "S2" ~ 12,
        TRUE ~ readr::parse_number(Irrigation_chr)
      )
    )

  if (diagnostics) {
    bad_irr <- irr_map %>% dplyr::filter(is.na(irr_num))
    if (nrow(bad_irr) > 0) {
      message("Unmapped Irrigation levels (these rows will be dropped): ",
              paste(bad_irr$Irrigation_chr, collapse = ", "))
    } else {
      message("All Irrigation levels mapped successfully (including S1/S2 if present).")
    }
  }

  d1 <- d0 %>%
    dplyr::mutate(Irrigation_chr = as.character(Irrigation)) %>%
    dplyr::left_join(irr_map, by = "Irrigation_chr")

  # ------------------------------------------------------------
  # Residue placeholder (Stan requires RES predictor)
  # Use an epsilon inside (0,1) as the boundary holder, never 0.
  # If a residue column exists, clamp it into (eps, 1-eps).
  # If missing entirely, create eps as a placeholder.
  # ------------------------------------------------------------
  eps_res <- 1e-4

  clamp01 <- function(x, eps = 1e-4) {
    x <- as.numeric(x)
    x[!is.finite(x)] <- NA_real_
    pmin(pmax(x, eps), 1 - eps)
  }

  if ("RES_z" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(RES_z), eps_res, clamp01(RES_z, eps_res)))
  } else if ("residue_z" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$residue_z), eps_res, clamp01(.data$residue_z, eps_res)))
  } else if ("RES_input" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$RES_input), eps_res, clamp01(.data$RES_input, eps_res)))
  } else if ("RES" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$RES), eps_res, clamp01(.data$RES, eps_res)))
  } else if ("residue_prop" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$residue_prop), eps_res, clamp01(.data$residue_prop, eps_res)))
  } else {
    d1 <- d1 %>% dplyr::mutate(RES_z = eps_res)
  }

  # ------------------------------------------------------------
  # Diagnostics BEFORE final dropping
  # ------------------------------------------------------------
  if (diagnostics) {

    na_irr_by_year <- d1 %>%
      dplyr::summarise(
        .by = Year,
        n = dplyr::n(),
        n_irr_na = sum(is.na(irr_num)),
        frac_irr_na = n_irr_na / n
      ) %>%
      dplyr::arrange(dplyr::desc(frac_irr_na))

    message("\n---- Diagnostic: NA irr_num by Year (top 10) ----")
    print(utils::head(na_irr_by_year, 10))

    diag_by_year <- d1 %>%
      dplyr::transmute(
        Year,
        na_event = is.na(.data[[event_id_col]]),
        na_stir  = is.na(Season_STIR_toDate),
        na_rep   = is.na(Rep),
        na_samp  = is.na(SampleMethod),
        na_flume = is.na(FlumeMethod),
        na_dup   = is.na(Duplicate),
        na_cin   = is.na(cin_z),
        na_crop  = is.na(Crop),
        na_res   = is.na(RES_z)
      ) %>%
      dplyr::summarise(
        .by = Year,
        n = dplyr::n(),
        frac_na_event = mean(na_event),
        frac_na_stir  = mean(na_stir),
        frac_na_rep   = mean(na_rep),
        frac_na_samp  = mean(na_samp),
        frac_na_flume = mean(na_flume),
        frac_na_dup   = mean(na_dup),
        frac_na_cin   = mean(na_cin),
        frac_na_crop  = mean(na_crop),
        frac_na_res   = mean(na_res)
      ) %>%
      dplyr::arrange(Year)

    message("\n---- Diagnostic: missing predictor fractions by Year ----")
    print(diag_by_year)
  }

  # ------------------------------------------------------------
  # Build final d_pred with model-ready columns
  # ------------------------------------------------------------
  d <- d1 %>%
    dplyr::transmute(
      event_id = .data[[event_id_col]],
      Year,
      Treatment,
      analyte_abbr,
      Crop,
      STIR_raw = Season_STIR_toDate,
      Rep,
      SampleMethod,
      FlumeMethod,
      irr_num,
      dup   = as.numeric(Duplicate),
      cin_z = if (cin_na_to_zero) dplyr::if_else(is.na(cin_z), 0, cin_z) else cin_z,
      RES_z = if (res_na_to_zero) dplyr::if_else(is.na(RES_z), 0, RES_z) else RES_z
    ) %>%
    dplyr::filter(
      !is.na(event_id),
      !is.na(Year),
      !is.na(Treatment),
      !is.na(analyte_abbr),
      !is.na(Crop),
      !is.na(STIR_raw),
      !is.na(Rep),
      !is.na(SampleMethod),
      !is.na(FlumeMethod),
      !is.na(irr_num),
      !is.na(dup),
      !is.na(cin_z),
      !is.na(RES_z)
    )

  if (diagnostics) {
    message("\n---- Final d_pred coverage by Year (after dropping incomplete rows) ----")
    print(d %>% dplyr::count(Year) %>% dplyr::arrange(Year))
  }

  d
}

# Run it
d_pred <- build_d_pred(wq_clean, event_id_col = "orig_row")

```


```{r real data generation}
make_annual_observed_table_all_treatments <- function(
  wq_clean,
  analyte_pick = NULL,
  keep_outflow_only = TRUE,
  drop_no_runoff = TRUE,
  B = 600,
  prob_ci = 0.95,
  seed = 1
) {
  needed <- c("Year","Treatment","analyte_abbr","Result_mg_L","Volume","orig_row")
  miss <- setdiff(needed, names(wq_clean))
  if (length(miss) > 0) stop("wq_clean missing columns: ", paste(miss, collapse = ", "))

  d0 <- wq_clean

  if (keep_outflow_only && "InflowOutflow" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(.data$InflowOutflow == "OUT")
  }
  if (drop_no_runoff && "NoRunoff" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(!isTRUE(.data$NoRunoff))
  }
  if (!is.null(analyte_pick)) {
    d0 <- d0 %>% dplyr::filter(as.character(.data$analyte_abbr) == analyte_pick)
  }

  # rows usable for observed load
  d_obs <- d0 %>%
    dplyr::filter(!is.na(.data$Result_mg_L), !is.na(.data$Volume)) %>%
    dplyr::mutate(
      Year = as.integer(.data$Year),
      analyte = as.character(.data$analyte_abbr),
      treatment = as.character(.data$Treatment),
      event_id = .data$orig_row,
      load_g = (.data$Result_mg_L * .data$Volume) / 1000,  # mg -> g
      vol_L  = .data$Volume
    ) %>%
    dplyr::select(Year, analyte, treatment, event_id, load_g, vol_L)

  q_lo <- (1 - prob_ci) / 2
  q_hi <- 1 - q_lo

  set.seed(seed)

  out <- d_obs %>%
    dplyr::group_by(.data$Year, .data$analyte, .data$treatment) %>%
    dplyr::group_modify(~{
      dat <- dplyr::distinct(.x, .data$event_id, .keep_all = TRUE)
      n_events <- nrow(dat)

      if (n_events == 0) {
        return(tibble::tibble(
          load_mean = NA_real_, load_low = NA_real_, load_high = NA_real_,
          volume_mean = NA_real_, volume_low = NA_real_, volume_high = NA_real_,
          conc_mean = NA_real_, conc_low = NA_real_, conc_high = NA_real_,
          n_events = 0L,
          prob_ci = prob_ci
        ))
      }

      # point estimates (annual totals)
      L_sum <- sum(dat$load_g, na.rm = TRUE)
      V_sum <- sum(dat$vol_L,  na.rm = TRUE)
      C_fw  <- if (is.finite(V_sum) && V_sum > 0) (L_sum * 1000 / V_sum) else NA_real_

      # If only one event, bootstrap is degenerate, CI == point estimate
      if (n_events == 1 || B <= 1) {
        return(tibble::tibble(
          load_mean = L_sum, load_low = L_sum, load_high = L_sum,
          volume_mean = V_sum, volume_low = V_sum, volume_high = V_sum,
          conc_mean = C_fw, conc_low = C_fw, conc_high = C_fw,
          n_events = as.integer(n_events),
          prob_ci = prob_ci
        ))
      }

      # bootstrap indices: n_events x B matrix
      idx_mat <- matrix(
        sample.int(n_events, size = n_events * B, replace = TRUE),
        nrow = n_events, ncol = B
      )

      load_mat <- matrix(dat$load_g[idx_mat], nrow = n_events, ncol = B)
      vol_mat  <- matrix(dat$vol_L[idx_mat],  nrow = n_events, ncol = B)

      Lb <- colSums(load_mat, na.rm = TRUE)
      Vb <- colSums(vol_mat,  na.rm = TRUE)
      Cb <- ifelse(Vb > 0, (Lb * 1000 / Vb), NA_real_)

      tibble::tibble(
        load_mean = L_sum,
        load_low  = stats::quantile(Lb, probs = q_lo, na.rm = TRUE, names = FALSE),
        load_high = stats::quantile(Lb, probs = q_hi, na.rm = TRUE, names = FALSE),

        volume_mean = V_sum,
        volume_low  = stats::quantile(Vb, probs = q_lo, na.rm = TRUE, names = FALSE),
        volume_high = stats::quantile(Vb, probs = q_hi, na.rm = TRUE, names = FALSE),

        conc_mean = C_fw,
        conc_low  = stats::quantile(Cb, probs = q_lo, na.rm = TRUE, names = FALSE),
        conc_high = stats::quantile(Cb, probs = q_hi, na.rm = TRUE, names = FALSE),

        n_events = as.integer(n_events),
        prob_ci = prob_ci
      )
    }) %>%
    dplyr::ungroup()

  out
}

# Example:
annual_tbl_obs <- make_annual_observed_table_all_treatments(
  wq_clean = wq_clean,
  B = 600,
  prob_ci = 0.95
)

```


```{r pred data table}
# ============================================================
# v1p7p5: Annual table from posterior predictions summed over d_pred
# IMPORTANT FIXES:
#   - De-duplicate to ONE row per event_id for each Year × Treatment × Analyte group before summing
#   - Irrigation is used on standardized scale irr_z (matches Stan data)
#   - beta_irr is analyte-specific in v1p7p5: post$beta_irr[draw, A]
# ============================================================
make_annual_load_table_all_treatments_pred <- function(
  post,
  d_pred,
  years_used,
  analyte_lookup,
  block_lookup,
  sampler_lookup,
  flume_lookup,
  crop_lookup,
  c_stats,
  V_mean, V_sd,
  stir_mean, stir_sd,
  irr_mean, irr_sd,             # NEW v1p7p5
  n_draws = 1500,
  prob_ci = 0.95,
  drop_unmapped_analytes = TRUE,
  drop_missing_c_stats = TRUE
) {
  needed_cols <- c(
    "event_id","Year","Treatment","analyte_abbr","STIR_raw",
    "Rep","SampleMethod","FlumeMethod","irr_num","dup","cin_z",
    "Crop","RES_z"
  )
  miss <- setdiff(needed_cols, names(d_pred))
  if (length(miss) > 0) stop("d_pred is missing columns: ", paste(miss, collapse = ", "))

  # draw subsample
  n_total <- nrow(post$alpha)
  draw_idx <- if (n_total > n_draws) sample.int(n_total, n_draws) else seq_len(n_total)
  D <- length(draw_idx)

  q_lo <- (1 - prob_ci) / 2
  q_hi <- 1 - q_lo
  ci <- function(x) stats::quantile(x, probs = c(q_lo, q_hi), na.rm = TRUE, names = FALSE)

  # map indices + stats
  dat <- d_pred %>%
    dplyr::left_join(analyte_lookup %>% dplyr::select(A, analyte_abbr), by = "analyte_abbr") %>%
    dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
    dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
    dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
    dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
    dplyr::left_join(c_stats %>% dplyr::select(A, c_mean, c_sd) %>% dplyr::distinct(), by = "A") %>%
    dplyr::mutate(
      Y      = match(Year, years_used),
      STIR_z = (STIR_raw - stir_mean) / stir_sd,
      irr_z  = (irr_num  - irr_mean)  / irr_sd     # NEW v1p7p5
    )

  # ---- drop unmapped analytes instead of stopping ----
  if (anyNA(dat$A)) {
    bad_an <- sort(unique(dat$analyte_abbr[is.na(dat$A)]))
    msg <- paste0(
      "Some analyte_abbr not found in analyte_lookup: ",
      paste(bad_an, collapse = ", "),
      "."
    )
    if (!isTRUE(drop_unmapped_analytes)) {
      stop(msg)
    } else {
      warning(msg, " Dropping these rows from d_pred for annual table.")
      dat <- dat %>% dplyr::filter(!is.na(A))
    }
  }

  # ---- optionally drop rows missing c_stats ----
  if (anyNA(dat$c_mean) || anyNA(dat$c_sd)) {
    bad_an2 <- sort(unique(dat$analyte_abbr[is.na(dat$c_mean) | is.na(dat$c_sd)]))
    msg2 <- paste0(
      "Missing c_stats (c_mean/c_sd) for some analytes: ",
      paste(bad_an2, collapse = ", "),
      "."
    )
    if (!isTRUE(drop_missing_c_stats)) {
      stop(msg2)
    } else {
      warning(msg2, " Dropping these rows from d_pred for annual table.")
      dat <- dat %>% dplyr::filter(!is.na(c_mean), !is.na(c_sd))
    }
  }

  # Hard checks for structural lookups (these *should* be complete)
  if (anyNA(dat$B))  stop("Some Rep not found in block_lookup.")
  if (anyNA(dat$S))  stop("Some SampleMethod not found in sampler_lookup.")
  if (anyNA(dat$Fu)) stop("Some FlumeMethod not found in flume_lookup.")
  if (anyNA(dat$Cr)) stop("Some Crop not found in crop_lookup.")
  if (anyNA(dat$Y))  stop("Some Year not found in years_used.")
  if (anyNA(dat$irr_z)) stop("Some irr_z are NA. Check irr_mean/irr_sd and irr_num mapping.")

  # Integer indices for fast indexing
  dat <- dat %>%
    dplyr::mutate(
      A  = as.integer(A),
      B  = as.integer(B),
      S  = as.integer(S),
      Fu = as.integer(Fu),
      Y  = as.integer(Y),
      Cr = as.integer(Cr)
    )

  # Groups: all treatments automatically
  groups <- dat %>%
    dplyr::distinct(Year, analyte_abbr, Treatment) %>%
    dplyr::arrange(analyte_abbr, Year, Treatment)

  out <- vector("list", nrow(groups))

  for (g in seq_len(nrow(groups))) {

    yr <- groups$Year[g]
    an <- groups$analyte_abbr[g]
    tr <- groups$Treatment[g]

    dsub <- dat %>% dplyr::filter(Year == yr, analyte_abbr == an, Treatment == tr)
    if (nrow(dsub) == 0) next

    # de-duplicate analyte rows to one per event_id
    dsub1 <- dsub %>% dplyr::distinct(event_id, .keep_all = TRUE)

    # Row-level predictors (use dsub1)
    A  <- dsub1$A;  B <- dsub1$B;  S <- dsub1$S;  Fu <- dsub1$Fu;  Y <- dsub1$Y;  Cr <- dsub1$Cr
    stz <- dsub1$STIR_z
    cin <- dsub1$cin_z
    irz <- dsub1$irr_z              # NEW v1p7p5
    dup <- dsub1$dup
    res <- dsub1$RES_z
    c_mean <- dsub1$c_mean
    c_sd   <- dsub1$c_sd

    # Event-level ledger for volumes (also 1 row per event_id)
    ev <- dsub %>%
      dplyr::distinct(event_id, Year, Treatment, STIR_z, Crop, RES_z, Rep, SampleMethod, FlumeMethod, irr_num, irr_z, dup) %>%
      dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
      dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
      dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
      dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
      dplyr::mutate(
        Cr = as.integer(Cr),
        B  = as.integer(B),
        S  = as.integer(S),
        Fu = as.integer(Fu)
      )

    stz_ev <- ev$STIR_z
    res_ev <- ev$RES_z
    Cr_ev  <- ev$Cr

    V_ann   <- numeric(D)
    L_ann_g <- numeric(D)
    C_fw    <- numeric(D)

    for (dd in seq_len(D)) {
      s <- draw_idx[dd]

      # ---- Predict volume per event (one per event_id) ----
      muV_z_ev <- post$a_V[s] +
        post$b_V[s]        * stz_ev +
        post$beta_res_V[s] * res_ev +
        post$gamma_Cr_V[s, Cr_ev]

      V_ev_raw <- V_mean + V_sd * muV_z_ev
      Vsum <- sum(V_ev_raw, na.rm = TRUE)

      # ---- Predict concentration per analyte event-row (one per event_id) ----
      muV_z_row <- post$a_V[s] +
        post$b_V[s]        * stz +
        post$beta_res_V[s] * res +
        post$gamma_Cr_V[s, Cr]

      gB <- post$gamma_B[s, , , drop = TRUE][cbind(A, B)]
      gS <- post$gamma_S[s, , , drop = TRUE][cbind(A, S)]
      gF <- post$gamma_F[s, , , drop = TRUE][cbind(A, Fu)]
      kY <- post$F_year[s, , , drop = TRUE][cbind(Y, A)]
      gCrC <- post$gamma_Cr[s, , , drop = TRUE][cbind(A, Cr)]

      muC_z <- post$alpha[s, A] +
        post$beta_stir[s, A]  * stz +
        post$beta_cin[s, A]   * cin +
        post$beta_vol[s]      * muV_z_row +
        post$beta_irr[s, A]   * irz +      # NEW v1p7p5: analyte-specific and standardized
        post$beta_dup[s, A]   * dup +
        post$beta_res_C[s, A] * res +
        gCrC +
        gB + gS + gF + kY

      C_raw <- c_mean + c_sd * muC_z

      # ---- Annual load for this analyte: sum across events ONCE ----
      V_row_raw <- V_mean + V_sd * muV_z_row
      Lmg <- sum(C_raw * V_row_raw, na.rm = TRUE)
      Lg  <- Lmg / 1000

      V_ann[dd]   <- Vsum
      L_ann_g[dd] <- Lg
      C_fw[dd]    <- if (is.finite(Vsum) && Vsum > 0) ((Lg * 1000) / Vsum) else NA_real_
    }

    out[[g]] <- tibble::tibble(
      Year = yr,
      analyte = an,
      treatment = tr,
      volume_mean   = mean(V_ann,   na.rm = TRUE),
      volume_low    = ci(V_ann)[1],
      volume_high   = ci(V_ann)[2],
      conc_mean     = mean(C_fw,    na.rm = TRUE),
      conc_low      = ci(C_fw)[1],
      conc_high     = ci(C_fw)[2],
      load_mean     = mean(L_ann_g, na.rm = TRUE),
      load_low      = ci(L_ann_g)[1],
      load_high     = ci(L_ann_g)[2],
      n_events      = nrow(ev),
      n_rows        = nrow(dsub1),
      prob_ci       = prob_ci
    )
  }

  dplyr::bind_rows(out) %>%
    dplyr::arrange(analyte, Year, treatment)
}

# Example usage:
annual_tbl_pred <- make_annual_load_table_all_treatments_pred(
  post = post,
  d_pred = d_pred,
  years_used = years_used,
  analyte_lookup = analyte_lookup,
  block_lookup = block_lookup,
  sampler_lookup = sampler_lookup,
  flume_lookup = flume_lookup,
  crop_lookup = crop_lookup,
  c_stats = c_stats,
  V_mean = V_mean, V_sd = V_sd,
  stir_mean = stir_mean, stir_sd = stir_sd,
  irr_mean = irr_mean, irr_sd = irr_sd
)

```


```{r plot annual loads fxn}
plot_annual_by_treatment_pretty <- function(
  annual_tbl,
  analyte_pick,
  metric = c("load","conc","volume"),
  source_label = NULL,
  draw_line = TRUE,
  show_points = TRUE,
  dodge = 0.20,
  prob_ci = NULL
) {
  metric <- match.arg(metric)

  stopifnot(all(c("Year","analyte","treatment") %in% names(annual_tbl)))

  y_mean <- paste0(metric, "_mean")
  y_low  <- paste0(metric, "_low")
  y_high <- paste0(metric, "_high")

  # Backward compatibility: accept old *_low95/*_high95
  if (!all(c(y_low, y_high) %in% names(annual_tbl))) {
    y_low  <- paste0(metric, "_low95")
    y_high <- paste0(metric, "_high95")
  }

  stopifnot(all(c(y_mean, y_low, y_high) %in% names(annual_tbl)))

  d <- annual_tbl %>%
    dplyr::filter(as.character(analyte) == analyte_pick) %>%
    dplyr::mutate(
      Year = as.integer(Year),
      treatment = factor(treatment, levels = c("CT","MT","ST"))
    ) %>%
    dplyr::arrange(Year, treatment)

  if (nrow(d) == 0) stop("No rows found for analyte = '", analyte_pick, "'.")

  # determine CI label
  if (is.null(prob_ci)) {
    if ("prob_ci" %in% names(d)) prob_ci <- unique(d$prob_ci)[1]
  }
  ci_txt <- if (!is.null(prob_ci) && is.finite(prob_ci)) {
    sprintf(" (%.0f%% interval)", 100 * prob_ci)
  } else {
    ""
  }

  ylab <- switch(
    metric,
    "load"   = paste0("Annual load (g)", ci_txt),
    "conc"   = paste0("Flow-weighted concentration (mg/L)", ci_txt),
    "volume" = paste0("Annual volume (L)", ci_txt)
  )

  pd <- ggplot2::position_dodge(width = dodge)

  # Okabe–Ito colorblind-safe palette
  trt_cols <- c(
    CT = "#0072B2",
    MT = "#E69F00",
    ST = "#009E73"
  )

  ttl <- paste0(analyte_pick, ": annual ", metric, " by treatment")
  if (!is.null(source_label)) ttl <- paste0(ttl, " (", source_label, ")")

  p <- ggplot2::ggplot(
    d,
    ggplot2::aes(
      x = Year,
      y = .data[[y_mean]],
      color = treatment,
      group = treatment
    )
  ) +
    ggplot2::geom_errorbar(
      ggplot2::aes(ymin = .data[[y_low]], ymax = .data[[y_high]]),
      width = 0,
      position = pd,
      linewidth = 1.6,
      alpha = 0.7
    )

  if (draw_line) {
    p <- p + ggplot2::geom_line(
      position = pd,
      linewidth = 1.1,
      alpha = 0.9
    )
  }

  if (show_points) {
    p <- p + ggplot2::geom_point(
      position = pd,
      size = 2.7,
      stroke = 0.2,
      shape = 16
    )
  }

  p +
    ggplot2::scale_color_manual(values = trt_cols) +
    ggplot2::scale_x_continuous(
      breaks = sort(unique(d$Year)),
      expand = ggplot2::expansion(mult = c(0.01, 0.02))
    ) +
    ggplot2::labs(
      x = "Year",
      y = ylab,
      title = ttl,
      color = "Treatment"
    ) +
    ggplot2::theme_classic(base_size = 14) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      plot.title = ggplot2::element_text(face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      legend.key.height = grid::unit(0.9, "lines"),
      panel.grid.major.y = ggplot2::element_line(color = "grey90", linewidth = 0.4)
    )
}
```

```{r plot annual loads execution}
# Modeled-only (v1p7p5 annual_tbl_pred)
p_mod <- plot_annual_by_treatment_pretty(
  annual_tbl = annual_tbl_pred,
  analyte_pick = "TSS",
  metric = "load",
  source_label = "Modeled",
  draw_line = TRUE
)

ggplot2::ggsave(
  filename = "../figs/TSS_annual_load_modeled_only_v1p7p5.png",
  plot = p_mod,
  width = 8,
  height = 5,
  dpi = 300
)

# Observed-only
p_obs <- plot_annual_by_treatment_pretty(
  annual_tbl = annual_tbl_obs,
  analyte_pick = "TSS",
  metric = "load",
  source_label = "Observed",
  draw_line = FALSE
)

ggplot2::ggsave(
  filename = "../figs/TSS_annual_load_observed_only_v1p7p5.png",
  plot = p_obs,
  width = 8,
  height = 5,
  dpi = 300
)

```


```{r bind obs and modeled}
bind_observed_modeled <- function(annual_tbl_pred, annual_tbl_obs) {

  # ----------------------------
  # Helper: normalize CI columns to *_low95/*_high95
  # ----------------------------
  normalize_ci <- function(df, prefix, source_label) {

    mean_nm <- paste0(prefix, "_mean")
    low95_nm  <- paste0(prefix, "_low95")
    high95_nm <- paste0(prefix, "_high95")
    low_nm    <- paste0(prefix, "_low")
    high_nm   <- paste0(prefix, "_high")

    if (all(c(mean_nm, low95_nm, high95_nm) %in% names(df))) return(df)

    if (all(c(mean_nm, low_nm, high_nm) %in% names(df))) {
      df[[low95_nm]]  <- df[[low_nm]]
      df[[high95_nm]] <- df[[high_nm]]
      return(df)
    }

    stop(
      sprintf(
        "%s table is missing CI columns for '%s'. Expected either (%s,%s,%s) or (%s,%s,%s). Present: %s",
        source_label, prefix,
        mean_nm, low95_nm, high95_nm,
        mean_nm, low_nm, high_nm,
        paste(names(df), collapse = ", ")
      )
    )
  }

  # ---- normalize both inputs ----
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "load",   "Modeled")
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "conc",   "Modeled")
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "volume", "Modeled")

  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "load",   "Observed")
  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "conc",   "Observed")
  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "volume", "Observed")

  # ----------------------------
  # Build n safely (non-vectorized column-existence checks)
  # ----------------------------
  if ("n_events" %in% names(annual_tbl_pred)) {
    annual_tbl_pred$n <- annual_tbl_pred$n_events
  } else if ("n_rows" %in% names(annual_tbl_pred)) {
    annual_tbl_pred$n <- annual_tbl_pred$n_rows
  } else {
    annual_tbl_pred$n <- NA_integer_
  }

  if ("n_events" %in% names(annual_tbl_obs)) {
    annual_tbl_obs$n <- annual_tbl_obs$n_events
  } else if ("n_rows" %in% names(annual_tbl_obs)) {
    annual_tbl_obs$n <- annual_tbl_obs$n_rows
  } else {
    annual_tbl_obs$n <- NA_integer_
  }

  # ---- modeled ----
  mod <- annual_tbl_pred %>%
    dplyr::transmute(
      Year      = as.integer(Year),
      analyte   = as.character(analyte),
      treatment = as.character(treatment),
      source    = "Modeled",

      load_mean,   load_low95,   load_high95,
      conc_mean,   conc_low95,   conc_high95,
      volume_mean, volume_low95, volume_high95,

      n = as.integer(n)
    )

  # ---- observed ----
  obs <- annual_tbl_obs %>%
    dplyr::transmute(
      Year      = as.integer(Year),
      analyte   = as.character(analyte),
      treatment = as.character(treatment),
      source    = "Observed",

      load_mean,   load_low95,   load_high95,
      conc_mean,   conc_low95,   conc_high95,
      volume_mean, volume_low95, volume_high95,

      n = as.integer(n)
    )

  dplyr::bind_rows(mod, obs) %>%
    dplyr::mutate(
      treatment = factor(treatment, levels = c("CT","MT","ST")),
      source    = factor(source, levels = c("Modeled","Observed"))
    ) %>%
    dplyr::arrange(analyte, Year, treatment, source)
}

# Combine modeled and observed annual tables
annual_both <- bind_observed_modeled(annual_tbl_pred, annual_tbl_obs)

```


```{r plot obs and mod}
plot_annual_by_treatment_obs_vs_mod <- function(
  annual_both,
  analyte_pick,
  metric = c("load","conc","volume"),
  dodge = 0.20
) {
  metric <- match.arg(metric)

  y_mean <- paste0(metric, "_mean")
  y_low  <- paste0(metric, "_low95")
  y_high <- paste0(metric, "_high95")

  stopifnot(all(c("Year","analyte","treatment","source", y_mean, y_low, y_high) %in% names(annual_both)))

  d <- annual_both %>%
    dplyr::filter(as.character(analyte) == analyte_pick) %>%
    dplyr::mutate(
      Year = as.integer(Year),
      treatment = factor(treatment, levels = c("CT","MT","ST")),
      source = factor(source, levels = c("Modeled","Observed"))
    ) %>%
    dplyr::arrange(Year, treatment, source)

  if (nrow(d) == 0) stop("No rows found for analyte = '", analyte_pick, "'.")

  ylab <- switch(
    metric,
    "load"   = "Annual load (g)",
    "conc"   = "Flow-weighted concentration (mg/L)",
    "volume" = "Annual volume (L)"
  )

  pd <- ggplot2::position_dodge(width = dodge)

  trt_cols <- c(
    CT = "#1f77b4",
    MT = "#ff7f0e",
    ST = "#2ca02c"
  )

  # Split for cleaner layering
  d_mod <- d %>% dplyr::filter(source == "Modeled")
  d_obs <- d %>% dplyr::filter(source == "Observed")

  ggplot2::ggplot() +
    # --- Modeled uncertainty (linerange) ---
    ggplot2::geom_linerange(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        ymin = .data[[y_low]],
        ymax = .data[[y_high]],
        color = treatment
      ),
      position = pd,
      linewidth = 1.6,
      alpha = 0.45,
      show.legend = FALSE
    ) +

    # --- Modeled line ---
    ggplot2::geom_line(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        color = treatment,
        group = treatment
      ),
      position = pd,
      linewidth = 1.2,
      alpha = 0.9
    ) +

    # --- Modeled points: filled with treatment color, black outline ---
    ggplot2::geom_point(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        fill = treatment
      ),
      position = pd,
      shape = 21,
      size = 3.4,
      stroke = 0.9,
      color = "black"
    ) +

    # --- Observed uncertainty (linerange) ---
    ggplot2::geom_linerange(
      data = d_obs,
      ggplot2::aes(
        x = Year,
        ymin = .data[[y_low]],
        ymax = .data[[y_high]],
        color = treatment
      ),
      position = pd,
      linewidth = 1.2,
      alpha = 0.35,
      show.legend = FALSE
    ) +

    # --- Observed points: white fill, treatment-colored outline ---
    ggplot2::geom_point(
      data = d_obs,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        color = treatment
      ),
      position = pd,
      shape = 21,
      fill = "white",
      size = 2.8,
      stroke = 1.2
    ) +

    ggplot2::scale_color_manual(values = trt_cols, name = "Treatment") +
    ggplot2::scale_fill_manual(values = trt_cols, guide = "none") +
    ggplot2::scale_x_continuous(
      breaks = sort(unique(d$Year)),
      expand = ggplot2::expansion(mult = c(0.01, 0.02))
    ) +
    ggplot2::labs(
      x = "Year",
      y = ylab,
      title = paste0(analyte_pick, ": annual ", metric, " by treatment"),
      shape = "Estimate type"
    ) +
    ggplot2::theme_bw(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      plot.title = ggplot2::element_text(face = "bold"),
      legend.position = "right",
      panel.grid.minor = ggplot2::element_blank()
    )
}

```

```{r execute plot both}
# Example usage:
plot_annual_by_treatment_obs_vs_mod(annual_both, analyte_pick = "NO3", metric = "load")
```

```{r save all plots}
# ------------------------------------------------------------
# Save images (v1p7p5)
# ------------------------------------------------------------
sanitize_slug <- function(x) {
  x <- as.character(x)
  x <- tolower(x)
  x <- gsub("[[:space:]]+", "_", x)
  x <- gsub("[^a-z0-9_]+", "", x)
  x <- gsub("_+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

save_all_analyte_obs_vs_mod_plots <- function(
  annual_both,
  metric = c("load","conc","volume"),
  out_dir = file.path("../figs", "annual_obs_vs_modeled_v1p7p5"),
  width = 9,
  height = 5.5,
  dpi = 300,
  device = "jpg"
) {
  metric <- match.arg(metric)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  analytes <- sort(unique(as.character(annual_both$analyte)))

  log <- vector("list", length(analytes))

  for (i in seq_along(analytes)) {
    a <- analytes[i]
    slug <- sanitize_slug(a)

    p <- try(
      plot_annual_by_treatment_obs_vs_mod(
        annual_both = annual_both,
        analyte_pick = a,
        metric = metric
      ),
      silent = TRUE
    )

    if (inherits(p, "try-error")) {
      log[[i]] <- tibble::tibble(
        analyte = a,
        saved = FALSE,
        file = NA_character_,
        error = as.character(p)
      )
      next
    }

    fname <- file.path(out_dir, paste0("annual_", metric, "_", slug, "_obs_vs_modeled_v1p7p5.", device))

    ggplot2::ggsave(
      filename = fname,
      plot = p,
      width = width,
      height = height,
      dpi = dpi,
      device = device
    )

    log[[i]] <- tibble::tibble(
      analyte = a,
      saved = TRUE,
      file = fname,
      error = NA_character_
    )
  }

  dplyr::bind_rows(log)
}

# Example:
log_load <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "load")
# log_conc <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "conc")
# log_vol  <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "volume")

```



```{r export bayes annual predictions}
# ============================================================
# v1p7p5: Export annual posterior predictions (and optional observed)
# ============================================================

export_dir <- "../out"
dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)

model_version <- "v1p7p5"

# ---- 1) Annual posterior summary (Modeled) ----
annual_pred_out <- annual_tbl_pred %>%
  dplyr::mutate(
    model_version = model_version,
    source = "Bayes_Modeled",
    interval_prob = dplyr::coalesce(prob_ci, 0.95)
  ) %>%
  dplyr::select(
    model_version, source, interval_prob,
    Year, analyte, treatment,
    n_events, n_rows,
    volume_mean, volume_low, volume_high,
    conc_mean,   conc_low,   conc_high,
    load_mean,   load_low,   load_high
  ) %>%
  dplyr::arrange(analyte, Year, treatment)

pred_file <- file.path(export_dir, paste0("annual_load_summary_bayes_", model_version, ".csv"))
readr::write_csv(annual_pred_out, pred_file)
message("[OK] Wrote: ", pred_file)

# ---- 2) Observed annual summary in same schema ----
pick_ci_cols <- function(df, prefix) {
  mean_nm  <- paste0(prefix, "_mean")
  low95_nm <- paste0(prefix, "_low95")
  high95_nm <- paste0(prefix, "_high95")
  low_nm   <- paste0(prefix, "_low")
  high_nm  <- paste0(prefix, "_high")

  if (all(c(mean_nm, low95_nm, high95_nm) %in% names(df))) {
    list(mean = mean_nm, low = low95_nm, high = high95_nm)
  } else if (all(c(mean_nm, low_nm, high_nm) %in% names(df))) {
    list(mean = mean_nm, low = low_nm, high = high_nm)
  } else {
    stop(
      sprintf(
        "Observed table missing CI columns for '%s'. Need (%s,%s,%s) or (%s,%s,%s). Present: %s",
        prefix,
        mean_nm, low95_nm, high95_nm,
        mean_nm, low_nm, high_nm,
        paste(names(df), collapse = ", ")
      )
    )
  }
}

vol_cols  <- pick_ci_cols(annual_tbl_obs, "volume")
conc_cols <- pick_ci_cols(annual_tbl_obs, "conc")
load_cols <- pick_ci_cols(annual_tbl_obs, "load")

annual_obs_out <- annual_tbl_obs %>%
  dplyr::mutate(
    model_version = model_version,
    source = "Observed",
    interval_prob = if ("prob_ci" %in% names(.)) dplyr::coalesce(prob_ci, 0.95) else 0.95
  ) %>%
  dplyr::transmute(
    model_version, source, interval_prob,
    Year,
    analyte   = as.character(analyte),
    treatment = as.character(treatment),
    n_events,
    n_rows = NA_integer_,

    volume_mean = .data[[vol_cols$mean]],
    volume_low  = .data[[vol_cols$low]],
    volume_high = .data[[vol_cols$high]],

    conc_mean = .data[[conc_cols$mean]],
    conc_low  = .data[[conc_cols$low]],
    conc_high = .data[[conc_cols$high]],

    load_mean = .data[[load_cols$mean]],
    load_low  = .data[[load_cols$low]],
    load_high = .data[[load_cols$high]]
  ) %>%
  dplyr::arrange(analyte, Year, treatment)

# ---- 3) Combined (Modeled + Observed) ----
annual_combined_out <- dplyr::bind_rows(annual_pred_out, annual_obs_out) %>%
  dplyr::arrange(source, analyte, Year, treatment)

comb_file <- file.path(export_dir, paste0("annual_load_summary_bayes_plus_observed_", model_version, ".csv"))
readr::write_csv(annual_combined_out, comb_file)
message("[OK] Wrote: ", comb_file)

# ---- 4) Optional: per-draw export (for Python fan charts / method-comparison) ----
export_annual_draws <- function(
  post,
  d_pred,
  years_used,
  analyte_lookup,
  block_lookup,
  sampler_lookup,
  flume_lookup,
  crop_lookup,
  c_stats,
  V_mean, V_sd,
  stir_mean, stir_sd,
  irr_mean, irr_sd,            # NEW v1p7p5
  n_draws = 400,
  seed = 1,
  drop_unmapped_analytes = TRUE,
  drop_missing_c_stats = TRUE,
  drop_unmapped_years = FALSE,
  drop_unmapped_other = FALSE
) {
  needed_cols <- c(
    "event_id","Year","Treatment","analyte_abbr","STIR_raw",
    "Rep","SampleMethod","FlumeMethod","irr_num","dup","cin_z",
    "Crop","RES_z"
  )
  miss <- setdiff(needed_cols, names(d_pred))
  if (length(miss) > 0) stop("d_pred is missing columns: ", paste(miss, collapse = ", "))

  set.seed(seed)

  # draw subsample
  n_total <- nrow(post$alpha)
  draw_idx <- if (n_total > n_draws) sample.int(n_total, n_draws) else seq_len(n_total)
  D <- length(draw_idx)

  # map indices + stats
  dat <- d_pred %>%
    dplyr::left_join(analyte_lookup %>% dplyr::select(A, analyte_abbr), by = "analyte_abbr") %>%
    dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
    dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
    dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
    dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
    dplyr::left_join(c_stats %>% dplyr::select(A, c_mean, c_sd) %>% dplyr::distinct(), by = "A") %>%
    dplyr::mutate(
      Y      = match(Year, years_used),
      STIR_z = (STIR_raw - stir_mean) / stir_sd,
      irr_z  = (irr_num  - irr_mean)  / irr_sd     # NEW v1p7p5
    )

  # ---- Diagnose + optionally drop unmapped analytes ----
  if (anyNA(dat$A)) {
    bad_an <- sort(unique(dat$analyte_abbr[is.na(dat$A)]))
    msg <- paste0(
      "Unmapped analyte_abbr (not in analyte_lookup): ",
      paste(bad_an, collapse = ", "), "."
    )
    if (!isTRUE(drop_unmapped_analytes)) stop(msg)
    warning(msg, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(A))
  }

  # ---- Diagnose + optionally drop missing c_stats ----
  if (anyNA(dat$c_mean) || anyNA(dat$c_sd)) {
    bad_an2 <- sort(unique(dat$analyte_abbr[is.na(dat$c_mean) | is.na(dat$c_sd)]))
    msg2 <- paste0(
      "Missing c_stats (c_mean/c_sd) for analytes: ",
      paste(bad_an2, collapse = ", "), "."
    )
    if (!isTRUE(drop_missing_c_stats)) stop(msg2)
    warning(msg2, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(c_mean), !is.na(c_sd))
  }

  # ---- Years_used mismatch (optional drop) ----
  if (anyNA(dat$Y)) {
    bad_y <- sort(unique(dat$Year[is.na(dat$Y)]))
    msgy <- paste0("Some Year not found in years_used: ", paste(bad_y, collapse = ", "), ".")
    if (!isTRUE(drop_unmapped_years)) stop(msgy)
    warning(msgy, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(Y))
  }

  # ---- Other lookups (optional drop; default: strict) ----
  if (anyNA(dat$B) || anyNA(dat$S) || anyNA(dat$Fu) || anyNA(dat$Cr)) {
    parts <- c()
    if (anyNA(dat$B))  parts <- c(parts, paste0("Rep: ", paste(sort(unique(dat$Rep[is.na(dat$B)])), collapse = ", ")))
    if (anyNA(dat$S))  parts <- c(parts, paste0("SampleMethod: ", paste(sort(unique(dat$SampleMethod[is.na(dat$S)])), collapse = ", ")))
    if (anyNA(dat$Fu)) parts <- c(parts, paste0("FlumeMethod: ", paste(sort(unique(dat$FlumeMethod[is.na(dat$Fu)])), collapse = ", ")))
    if (anyNA(dat$Cr)) parts <- c(parts, paste0("Crop: ", paste(sort(unique(dat$Crop[is.na(dat$Cr)])), collapse = ", ")))
    msgo <- paste0("Index mapping produced NAs for: ", paste(parts, collapse = " | "), ".")

    if (!isTRUE(drop_unmapped_other)) stop(msgo)
    warning(msgo, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(B), !is.na(S), !is.na(Fu), !is.na(Cr))
  }

  if (anyNA(dat$irr_z)) stop("Some irr_z are NA. Check irr_mean/irr_sd and irr_num mapping.")

  # Integer indices
  dat <- dat %>%
    dplyr::mutate(
      A  = as.integer(A),
      B  = as.integer(B),
      S  = as.integer(S),
      Fu = as.integer(Fu),
      Y  = as.integer(Y),
      Cr = as.integer(Cr)
    )

  groups <- dat %>%
    dplyr::distinct(Year, analyte_abbr, Treatment) %>%
    dplyr::arrange(analyte_abbr, Year, Treatment)

  out <- vector("list", length = D * nrow(groups))
  kk <- 1L

  for (dd in seq_len(D)) {
    s <- draw_idx[dd]

    for (g in seq_len(nrow(groups))) {

      yr <- groups$Year[g]
      an <- groups$analyte_abbr[g]
      tr <- groups$Treatment[g]

      dsub <- dat %>% dplyr::filter(Year == yr, analyte_abbr == an, Treatment == tr)
      if (nrow(dsub) == 0) next

      # de-duplicate to one row per event_id for analyte
      dsub1 <- dsub %>% dplyr::distinct(event_id, .keep_all = TRUE)

      A  <- dsub1$A;  B <- dsub1$B;  Sidx <- dsub1$S;  Fu <- dsub1$Fu;  Y <- dsub1$Y;  Cr <- dsub1$Cr
      stz <- dsub1$STIR_z
      cin <- dsub1$cin_z
      irz <- dsub1$irr_z               # NEW v1p7p5
      dup <- dsub1$dup
      res <- dsub1$RES_z
      c_mean <- dsub1$c_mean
      c_sd   <- dsub1$c_sd

      ev <- dsub %>%
        dplyr::distinct(event_id, STIR_z, RES_z, Crop) %>%
        dplyr::left_join(crop_lookup %>% dplyr::select(Cr, Crop), by = "Crop") %>%
        dplyr::mutate(Cr = as.integer(Cr)) %>%
        dplyr::filter(!is.na(Cr))

      muV_z_ev <- post$a_V[s] +
        post$b_V[s]         * ev$STIR_z +
        post$beta_res_V[s]  * ev$RES_z +
        post$gamma_Cr_V[s, ev$Cr]

      Vsum <- sum(V_mean + V_sd * muV_z_ev, na.rm = TRUE)

      muV_z_row <- post$a_V[s] +
        post$b_V[s]         * stz +
        post$beta_res_V[s]  * res +
        post$gamma_Cr_V[s, Cr]

      gB <- post$gamma_B[s, , , drop = TRUE][cbind(A, B)]
      gS <- post$gamma_S[s, , , drop = TRUE][cbind(A, Sidx)]
      gF <- post$gamma_F[s, , , drop = TRUE][cbind(A, Fu)]
      kY <- post$F_year[s, , , drop = TRUE][cbind(Y, A)]
      gCrC <- post$gamma_Cr[s, , , drop = TRUE][cbind(A, Cr)]

      muC_z <- post$alpha[s, A] +
        post$beta_stir[s, A]  * stz +
        post$beta_cin[s, A]   * cin +
        post$beta_vol[s]      * muV_z_row +
        post$beta_irr[s, A]   * irz +      # NEW v1p7p5
        post$beta_dup[s, A]   * dup +
        post$beta_res_C[s, A] * res +
        gCrC + gB + gS + gF + kY

      C_raw <- c_mean + c_sd * muC_z
      V_row_raw <- V_mean + V_sd * muV_z_row

      Lmg <- sum(C_raw * V_row_raw, na.rm = TRUE)
      Lg  <- Lmg / 1000
      Cfw <- if (is.finite(Vsum) && Vsum > 0) ((Lg * 1000) / Vsum) else NA_real_

      out[[kk]] <- tibble::tibble(
        draw = dd,
        draw_id = s,
        Year = yr,
        analyte = an,
        treatment = tr,
        volume_L = Vsum,
        conc_fw_mg_L = Cfw,
        load_g = Lg
      )
      kk <- kk + 1L
    }
  }

  dplyr::bind_rows(out[seq_len(kk - 1L)]) %>%
    dplyr::arrange(analyte, Year, treatment, draw)
}

# Uncomment to write per-draw table (can be big):
annual_draws <- export_annual_draws(
  post = post,
  d_pred = d_pred,
  years_used = years_used,
  analyte_lookup = analyte_lookup,
  block_lookup = block_lookup,
  sampler_lookup = sampler_lookup,
  flume_lookup = flume_lookup,
  crop_lookup = crop_lookup,
  c_stats = c_stats,
  V_mean = V_mean, V_sd = V_sd,
  stir_mean = stir_mean, stir_sd = stir_sd,
  irr_mean = irr_mean, irr_sd = irr_sd,
  n_draws = 400,
  seed = 1
)

export_dir <- file.path(getwd(), "out")
dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

draws_file <- file.path(
  export_dir,
  paste0("annual_load_draws_bayes_", model_version, ".csv")
)

readr::write_csv(annual_draws, draws_file)

```
