---
title: "Bayesian Modeling of STIR Effects on Water Quality"
author: "AJ Brown"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Load backend cleaning + functions
getwd()
source("./stir-bayes-backend.R")
library(rethinking)
library(dplyr)
library(ggplot2)
```

# Model 1.0 
- STIR total effect on OUT concentration 
- with analyte-specific slopes and intercepts
- with other DAG-identified adjustment variables
-- { Analyte, Block, Inflow Conc., Irrigation Count, Outflow vol, Sampler Method, duplicate status, flume type }

# 1. Load and clean STIR–WQ dataset

```{r load-clean}
# Step 1: load raw merged dataset
wq_raw <- load_wq_stir(
  path = "out/wq_with_stir_by_season.csv",
  year_max = Inf
)

# Step 2: clean, type-enforce, standardize (ALL done in backend)
wq_clean <- clean_wq_stir(wq_raw)

# Add original row number for tracking
wq_clean <- wq_clean %>%
  mutate(orig_row = row_number())


glimpse(wq_clean)
```

---

# 2. Prepare model-ready dataset

You now choose which variables enter your model from the fields already standardized in the backend:

* `cout_z` = per-analyte standardized OUT concentrations
* `cin_z`  = per-analyte standardized INFLOW concentrations
* `stir_season_z` = standardized seasonal STIR
* `stir_cumall_z` = standardized cumulative STIR
* `analyte_abbr` = clean factor index

Select analytes for modeling:

```{r subset-analytes}
# analytes: OP   TP   ICP  NO3  TN   TSS  TKN  TSP  NH4  NPOC Se   NO2  NOx  TDS
analytes_keep <- c("TP", "TSS")   # modify as needed
# fullset for ease of use
analytes_keep <- c("OP", "TP", "ICP", "NO3", "TN", "TSS", 
                   "TKN", "TSP", "NH4", "NPOC", "Se", "NO2", 
                   "NOx", "TDS")

d_mod <- wq_clean %>%
  filter(analyte_abbr %in% analytes_keep) %>%
  mutate(mod_row = row_number()) %>%
  droplevels()

d_mod %>% count(analyte_abbr)
```

Create analyte index and build data list for ulam:

```{r make-stan-dat}
# 2. Build indices and standardized predictors -----------------------------

d_mod <- d_mod %>%
  mutate(
    # analyte index
    A = as.integer(analyte_abbr),

    # Block (Rep) index
    B = as.integer(Rep),

    # Sampler method index
    S = as.integer(SampleMethod),

    # Flume type index
    Fu = as.integer(FlumeMethod),

    # Irrigation count as numeric
    irr_num = as.integer(Irrigation),

    # Inflow concentration, already standardized in cin_z
    # For now, replace missing with 0 on the z scale so the linear term
    # drops out when there is no inflow measurement
    cin_std = if_else(is.na(cin_z), 0, cin_z),

    # Duplicate status as 0/1
    dup = as.integer(Duplicate)
  )

# prepare data list for stan
stan_dat <- list(
  N    = nrow(d_mod),
  J    = length(unique(d_mod$A)),     # analytes
  B_n  = length(unique(d_mod$B)),     # blocks
  S_n  = length(unique(d_mod$S)),     # sampler methods
  F_n  = length(unique(d_mod$Fu)),    # flume types

  # indices
  A    = d_mod$A,
  B    = d_mod$B,
  S    = d_mod$S,
  Fu    = d_mod$Fu,

  # outcome
  C    = d_mod$cout_z,                # standardized outflow concentration

  # main exposure
  STIR = d_mod$stir_season_z,         # or stir_cumall_z if you want

  # covariates from DAG:
  CIN   = d_mod$cin_std,              # inflow concentration (z)
  VOL   = d_mod$volume_z,             # outflow volume (z)
  IRR   = d_mod$irr_num,              # irrigation count (int)
  DUP   = d_mod$dup                   # duplicate status (0/1)
)
```

```{r}
# Identify the stan_dat entries that are vectors of length N (i.e., row-level variables)
vec_names <- names(stan_dat)[sapply(stan_dat, function(x) length(x) == stan_dat$N)]

# Loop through these and report NAs
for (nm in vec_names) {
  vals <- stan_dat[[nm]]
  na_idx <- which(is.na(vals))

  cat("\n==============================\n")
  cat("Variable:", nm, "\n")

  if (length(na_idx) == 0) {
    cat("No NA values detected.\n")
  } else {
    cat("NA count:", length(na_idx), "\n")
    cat("Rows with NA:", paste(head(na_idx, 100), collapse = ", "),
        if (length(na_idx) > 100) "..." else "", "\n\n")

    # Show the corresponding d_mod rows
    cat("Example offending rows from d_mod:\n")
    print(d_mod[na_idx[1:min(10, length(na_idx))], ])
  }
}

```




```{r}
# lookup code for later
analyte_lookup <- d_mod %>%
  distinct(A, analyte_abbr) %>%
  arrange(A)

block_lookup <- d_mod %>%
  distinct(B, Rep) %>%
  arrange(B)

sampler_lookup <- d_mod %>%
  distinct(S, SampleMethod) %>%
  arrange(S)

flume_lookup <- d_mod %>%
  distinct(Fu, FlumeMethod) %>%
  arrange(Fu)

```


# 3. Hierarchical Bayesian model

```{r fit-model, eval=T}
m_stir <- ulam(
  alist(
    # likelihood
    C ~ dnorm(mu, sigma),

    # linear predictor
    mu <- alpha[A] +
          beta_stir[A] * STIR +
          beta_cin[A]  * CIN +
          beta_vol  * VOL +
          beta_irr  * IRR +
          beta_dup  * DUP +
          gamma_B[B] +
          gamma_S[S] +
          gamma_F[Fu],

    # analyte level variation
    alpha[A]     ~ dnorm(a_bar, sigma_alpha),
    beta_stir[A] ~ dnorm(b_bar, sigma_beta),
    beta_cin[A]  ~ dnorm(c_bar, sigma_c),

    # block, sampler, flume effects
    gamma_B[B] ~ dnorm(0, sigma_B),
    gamma_S[S] ~ dnorm(0, sigma_S),
    gamma_F[Fu] ~ dnorm(0, sigma_Fu),

    # population level means and slopes
    c(a_bar, b_bar, c_bar) ~ dnorm(0, 1),
    beta_vol  ~ dnorm(0, 1),
    beta_irr  ~ dnorm(0, 1),
    beta_dup  ~ dnorm(0, 1),
    
    # priors for imputing missing data (i.e., NAs)
    VOL ~ dnorm(0, 1), # When VOl observed, likelihood contribution. When NA, prior contribution.

    # scale parameters
    sigma_alpha ~ dexp(1),
    sigma_beta  ~ dexp(1),
    sigma_c     ~ dexp(1),
    sigma_B     ~ dexp(1),
    sigma_S     ~ dexp(1),
    sigma_Fu     ~ dexp(1),
    sigma       ~ dexp(1)
  ),
  data  = stan_dat,
  chains = 4,
  cores  = 4,
  iter   = 2000
)

```

---

# 4. Posterior inspection

```{r summarize, eval=FALSE}
#traceplot(m_stir)
#trankplot(m_stir)
precis(m_stir, depth=2, prob=0.95)
plot(precis(m_stir, depth=1, prob=0.95))
```

```{r extract, eval=FALSE}
# extract posterior samples
post <- extract.samples(m_stir)

# analyte-specific STIR slopes
alpha_hat      <- apply(post$alpha,     2, mean)
beta_stir_mean <- apply(post$beta_stir, 2, mean)

# 89 percent HPDI and 95 percent PI for beta_stir
beta_stir_hpdi <- apply(post$beta_stir, 2, HPDI, prob = 0.89)
beta_stir_pi   <- apply(post$beta_stir, 2, PI,   prob = 0.95)

analyte_lookup <- d_mod %>% 
  distinct(A, analyte_abbr) %>%
  arrange(A)

analyte_effects <- analyte_lookup %>%
  mutate(
    beta_stir_mean   = beta_stir_mean,
    beta_stir_median = apply(post$beta_stir, 2, median),
    beta_hpdi_low    = beta_stir_hpdi[1, ],
    beta_hpdi_high   = beta_stir_hpdi[2, ],
    beta_pi_low      = beta_stir_pi[1, ],
    beta_pi_high     = beta_stir_pi[2, ]
  )

analyte_effects

```

---

# 5. Plot posterior slopes

```{r plot-slopes, eval=FALSE}
# Recreate an index for plotting
A_seq        <- analyte_effects$A
beta_mean    <- analyte_effects$beta_stir_mean
beta_hpdi    <- analyte_effects %>% select(beta_hpdi_low, beta_hpdi_high)
beta_pi      <- analyte_effects %>% select(beta_pi_low, beta_pi_high)

# Base R plot setup
plot(
  A_seq, beta_mean,
  xaxt = "n",
  pch  = 16,
  col  = "darkred",
  ylim = range(c(beta_pi$beta_pi_low, beta_pi$beta_pi_high)),
  ylab = "Effect of STIR on Concentration",
  xlab = "Analyte",
  main = "Analyte-specific STIR effects with 89% (Darker) 95% (Lighter) CIs"
)

axis(1, at = A_seq, labels = analyte_effects$analyte_abbr, las = 2)

# 95 percent PI vertical bands (more transparent)
for (i in seq_along(A_seq)) {
  lines(
    x = c(A_seq[i], A_seq[i]),
    y = c(beta_pi$beta_pi_low[i], beta_pi$beta_pi_high[i]),
    col = col.alpha("darkred", 0.2),
    lwd = 5
  )
}

# 89 percent HPDI vertical bands
for (i in seq_along(A_seq)) {
  lines(
    x = c(A_seq[i], A_seq[i]),
    y = c(beta_hpdi$beta_hpdi_low[i], beta_hpdi$beta_hpdi_high[i]),
    col = col.alpha("darkred", 0.4),
    lwd = 5
  )
}

# Zero reference line
abline(h = 0, lty = 2)


```
