---
title: "Bayesian Modeling of STIR Effects on Water Quality"
author: "AJ Brown"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Load backend cleaning + functions
getwd()
source("./stir-bayes-backend.R")
library(rethinking)
library(dplyr)
library(ggplot2)
```

# 1. Load and clean STIR–WQ dataset

```{r load-clean}
# Step 1: load raw merged dataset
wq_raw <- load_wq_stir(
  path = "out/wq_with_stir_by_season.csv",
  year_max = Inf
)

# Step 2: clean, type-enforce, standardize (ALL done in backend)
wq_clean <- clean_wq_stir(wq_raw)

glimpse(wq_clean)
```

---

# 2. Prepare model-ready dataset

You now choose which variables enter your model from the fields already standardized in the backend:

* `cout_z` = per-analyte standardized OUT concentrations
* `cin_z`  = per-analyte standardized INFLOW concentrations
* `stir_season_z` = standardized seasonal STIR
* `stir_cumall_z` = standardized cumulative STIR
* `analyte_abbr` = clean factor index

Select analytes for modeling:

```{r subset-analytes}
analytes_keep <- c("TP", "TSS")   # modify as needed

d_mod <- wq_clean %>%
  filter(analyte_abbr %in% analytes_keep)

d_mod %>% count(analyte_abbr)
```

Create analyte index and build data list for ulam:

```{r make-stan-dat}
d_mod <- d_mod %>% 
  mutate(
    A = as.integer(analyte_abbr)
  )

stan_dat <- list(
  N    = nrow(d_mod),
  J    = length(unique(d_mod$A)),
  A    = as.factor(d_mod$A),
  C    = d_mod$cout_z,          # standardized concentration
  STIR = d_mod$stir_season_z    # standardized STIR predictor
)

str(stan_dat)
```

---

# 3. Hierarchical Bayesian model

```{r fit-model, eval=T}
m_stir <- ulam(
  alist(
    C ~ dnorm(mu, sigma),
    mu <- alpha[A] + beta[A] * STIR,

    alpha[A] ~ dnorm(a_bar, sigma_alpha),
    beta[A]  ~ dnorm(b_bar, sigma_beta),

    c(a_bar, b_bar) ~ dnorm(0, 1),
    sigma_alpha ~ dexp(1),
    sigma_beta  ~ dexp(1),
    sigma       ~ dexp(1)
  ),
  data = stan_dat,
  chains = 4,
  cores = 4,
  iter = 2000
)
```

---

# 4. Posterior inspection

```{r summarize, eval=FALSE}
precis(m_stir, depth=2)
plot(precis(m_stir, depth=2))
```

```{r extract, eval=FALSE}
post <- extract.samples(m_stir)

alpha_hat <- apply(post$alpha, 2, mean)
beta_hat  <- apply(post$beta,  2, mean)

analyte_lookup <- d_mod %>% 
  distinct(A, analyte_abbr) %>%
  arrange(A)

analyte_effects <- analyte_lookup %>%
  mutate(
    alpha_hat = alpha_hat,
    beta_hat  = beta_hat
  )

analyte_effects
```

---

# 5. Plot posterior slopes

```{r plot-slopes, eval=FALSE}
ggplot(analyte_effects, aes(x = analyte_abbr, y = beta_hat)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point() +
  labs(
    x = "Analyte",
    y = "Posterior mean slope of C_z on STIR_z",
    title = "Analyte-specific STIR effects on standardized concentration"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 6. Next steps

* Add censoring models for nondetects
* Compare seasonal STIR vs cumulative STIR
* Include treatment-level or year-level random effects
* Fit analyte-specific models for sensitivity analysis

